\documentclass{article} % For LaTeX2e
\usepackage{iclr2026_conference,times}

% Optional math commands from https://github.com/goodfeli/dlbook_notation.
\input{math_commands.tex}

\usepackage{hyperref}
\usepackage{url}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,fit,calc}
\usepackage{graphicx}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}


\title{\textbf{The Art of Agentic Systems Programming} \\ \large Chapter 3: Hardware-Aware Agent Training}
\author{Jiacheng~Y. \\
\texttt{github.com/dylanyunlon/operatorRL} \\
\texttt{github.com/microsoft/agent-lightning} \\
\texttt{github.com/dylanyunlon/theArtofPerformence} \\[6pt]
{\small \textit{Structured after D.\,E.\,Knuth, \textbf{The Art of Computer Programming}, Vol.\,2, Ch.\,3}}}
\date{}

% Authors must not appear in the submitted version. They should be hidden
% as long as the \iclrfinalcopy macro remains commented out below.
% Non-anonymous submissions will be rejected without review.


% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to \LaTeX{} to determine where to break
% the lines. Using \AND forces a linebreak at that point. So, if \LaTeX{}
% puts 3 of 4 authors names on the first line, and the last on the second
% line, try using \AND instead of \And before the third author name.

\newcommand{\fix}{\marginpar{FIX}}
\newcommand{\new}{\marginpar{NEW}}

%\iclrfinalcopy % Uncomment for camera-ready version, but NOT for submission.
\begin{document}


\maketitle

\begin{abstract}
This chapter develops the theory and practice of hardware-aware agent training by transposing the structure of Knuth's analysis of random number generation (TAOCP Vol.~2, Ch.~3) onto the domain of agentic reinforcement learning systems with heterogeneous accelerator backends. \S\ref{sec:3.1} introduces the problem: why assembly-level kernel optimization is necessary for self-deploying, self-evolving agent systems. \S\ref{sec:3.2} develops the mathematical framework of configuration periodicity and \S\ref{sec:3.2.hw} constructs the migration map from NVIDIA's Megatron-LM ecosystem to AWS Trainium2, analyzing each layer---parallelism primitives, fused kernels, communication overlap, memory management---and identifying the CCCL/DIA project's position in the NVIDIA fortress. \S\ref{sec:3.3} provides validation tests (chi-square, run test, spectral test, credit assignment) for assessing agentic training quality on heterogeneous hardware. The technical substrate is drawn from two open-source projects: \texttt{operatorRL} (kernel-level agent governance) and \texttt{agent-lightning} (RL training framework with VERL backend).
\end{abstract}


%% ================================================================
%% SECTION 3.1 — INTRODUCTION
%% Mapped from: 3.1.txt (TAOCP §3.1 Introduction to Random Numbers)
%% ================================================================
\section{Introduction}
\label{sec:3.1}

``Autonomously optimized'' agent training kernels have many different types of applications. For example:

\medskip

\noindent\textbf{a) Self-Deployment.} When using a cluster of Trainium2 accelerators to perform autonomous agent rollouts, it is necessary to use assembly-level kernel optimization to make throughput viable. Self-deployment involves domains ranging from distributed reinforcement learning (where gradient updates are subject to hardware-specific memory hierarchies) all the way to production serving (for instance, agents autonomously entering a live tool-use environment at unpredictable intervals).

\medskip

\noindent\textbf{b) Environment Feedback.} Usually, it is impractical to examine all possible agent trajectories, and hardware-aware trajectory sampling will enable us to understand ``typical'' reward landscapes.

\medskip

\noindent\textbf{c) Numerical Kernel Synthesis.} Using hand-written Trainium assembly, many ingenious techniques have been devised to solve the core computational primitives---matrix multiplication, attention, and collective communication. There are already considerable specialized works on this topic.

\medskip

\noindent\textbf{d) Agent Systems Programming.} To verify the effectiveness of agentic RL algorithms, hardware-optimized rollout engines are an excellent source of throughput---more importantly, hardware-optimized kernels are essential for the operation of self-evolving training loops, which frequently surpass deterministic optimization pipelines by a wide margin. The use of assembly-level hardware optimization is the principal application we are concerned with throughout this book, so we shall treat it in Chapter~3, before discussing most other agent training algorithms.

\medskip

\noindent\textbf{e) Orchestration Decisions.} It is said that many orchestration schedulers make their resource allocation decisions by round-robin or heuristic load balancing, and the like. It is also reported that certain cloud providers use similar bases for pricing. Sometimes, making such completely ``hardware-unaware'' decisions is acceptable. But hardware-aware scheduling is also an indispensable part of optimal multi-tenant agent serving strategies.

\medskip

\noindent\textbf{f) Expressiveness.} A degree of hardware-software co-design makes agent-generated outputs seem more capable. For example, the architecture below on the left---where kernels are hand-tuned per hardware target---is in many situations more performant than the one on the right, which relies entirely on a vendor compiler.

\medskip

See the Agent Lightning architecture: \texttt{agentlightning/trainer}, \texttt{agentlightning/runner}, \texttt{agentlightning/store}.

\medskip

\noindent\textbf{g) Benchmarking.} Running standardized agent tasks, measuring reward convergence, comparing kernel throughput---these are activities people value greatly. These traditional uses of hardware-aware agent training have been named the ``Operator RL method,'' and this has become a general term describing any training algorithm that couples agent rollout with direct hardware kernel optimization.

\bigskip

People who consider this topic almost always encounter a philosophical discussion about the meaning of ``hardware independence.'' In a certain sense, no truly hardware-independent agent training stack exists. For example, is a JAX-compiled XLA graph hardware-independent? We would rather speak of \emph{a portable sequence of hardware-optimized kernels with a deterministic compilation target}. This roughly means that each kernel's behavior is optimized for its specific accelerator, independently of how other kernels in the stack are compiled, and each kernel achieves a definite fraction of peak hardware utilization.

In a uniform hardware abstraction across accelerator targets, the CUDA runtime, the TPU runtime, and the Trainium Neuron runtime each present an interface where each possible kernel has an equal probability of achieving optimal utilization. All abstractions are generally assumed to be uniform, unless a specific hardware constraint is noted.

In a multi-backend deployment, the target platforms---NVIDIA GPU via CUDA, Google TPU via XLA, and AWS Trainium via Neuron/assembly---carry \emph{asymmetric} deployment risk. NVIDIA's CUDA ecosystem has over a decade of maturity, millions of developers, and comprehensive tooling; its deployment risk is lowest. Google's TPU/XLA ecosystem is mature for internal Google workloads but has a smaller external developer base. AWS Trainium's ecosystem is youngest: the Neuron SDK has iterated through 27+ major versions with breaking changes, the NKI programming model was introduced only with Neuron SDK 2.21, and the developer community is smallest. A realistic risk partition might assign approximately 10\% to GPU (mature but vendor-locked), 30\% to TPU (mature internally but external tooling gaps), and 60\% to Trainium (youngest ecosystem, highest technical uncertainty, but greatest strategic optionality). The engineering cost of maintaining all three is not $3\times$ a single platform but closer to $3 \cdot k \cdot d$ in the worst case (\S\ref{sec:3.2.critical}), because the debug spaces are largely disjoint.

Thus, if we are portably compiling a kernel for three backends, and the first two happen to achieve 95\% hardware utilization on GPU and TPU respectively, the probability that the third (Trainium) also achieves 95\% utilization is \emph{not} determined by the hardware's peak FLOPS and memory bandwidth alone---those set only an upper bound. The actual utilization depends on kernel optimization quality: how well the compiler (or the human kernel engineer) maps the computation onto the specific hardware's execution model. Peak FLOPS tells you the ceiling; kernel engineering determines how close you get. This distinction---between hardware \emph{capability} and kernel \emph{quality}---is precisely why Anthropic invests in hand-written Trainium assembly rather than relying on the Neuron compiler's auto-generated kernels.

There are many good ways to give an abstract definition of hardware portability for agent training stacks. We shall return to this interesting question in Section~3.5. But for now, let us understand the concept intuitively!

Many years ago, those who needed high-performance agent training in their research obtained it by carefully hand-tuning CUDA kernels for a single GPU architecture---extracting performance from one specific hardware configuration at a time. In 2018, the NVIDIA Megatron-LM framework \citep{shoeybi2019megatron} published methods for distributed training across thousands of GPUs, establishing the dominant paradigm for large-scale model training. Since then, several specialized frameworks have been built to extend this approach. In 2020, Google's approach with TPU pods and the JAX/XLA compilation stack offered an alternative path, compiling high-level operations down to TPU-optimized instructions; this capability had been championed internally at Google Brain for years. In 2023, AWS's Trainium chips, designed by Annapurna Labs, first became available at scale with an open ISA that allowed advanced users to write assembly-level kernels directly---bypassing the Neuron compiler's intermediate representations. In 2024, AWS released Trainium2 with substantially improved compute density, and Anthropic became one of the first organizations to write production assembly for these chips. A celebrated hardware-software co-design effort between Anthropic engineers and Annapurna chip architects has been ongoing for several years, producing optimized kernels for Claude's training and inference.

[See the Semianalysis report \citep{semianalysis2024trainium2} on Trainium2 architecture and Anthropic's kernel engineering approach.]

\bigskip

Shortly after the Megatron-LM framework demonstrated that distributed GPU training could scale to thousands of devices, people began to explore efficient methods for performing agent training across heterogeneous hardware. One could use CUDA as the sole backend, but due to vendor lock-in and pricing constraints, this approach's practicality is limited, because such a stack may be too expensive, and maintaining dependence on a single vendor is a troublesome affair. One could connect specialized hardware like TPUs or Trainium chips to the training loop, as Google and AWS have done, but this is also not entirely satisfactory, because when debugging a multi-platform kernel stack, it is practically impossible to reproduce the exact same numerical behavior across backends; moreover, each platform's toolchain may develop hard-to-detect regressions. However, developments in the 2020s have made the multi-platform approach viable again, because open ISAs and assembly-level access can distribute tens of gigabytes of optimized kernel code across backends. In 2024, Anthropic's engineers achieved direct Trainium assembly optimization for production workloads (combining deterministic kernel behavior with aggressive hardware-specific tuning), thus advancing the revival of multi-platform training stacks.

The disadvantages of the early single-vendor approach to agent training hardware led to interest in using a framework's own compiler infrastructure to produce portable kernels. Around 2019, the XLA (Accelerated Linear Algebra) compiler was among the first to propose this approach; its method was to take a high-level graph representation of the computation and lower it through successive optimization passes to target-specific code. For example, if the system is generating optimized GEMM kernels, the previous kernel's tiling configuration is taken, and the compiler searches for an optimal tiling for the new target---thus the next kernel configuration depends on the previous one.

There is a quite obvious objection to this technique: since each kernel's configuration is completely determined by the compiler's cost model, how can the sequence of kernels so produced be truly optimal? (See the remarks by Anthropic's engineering team at the beginning of this chapter.) The answer is that this sequence is not optimal; it merely \emph{appears} optimal. In typical applications, the true relationship between one kernel's configuration and the hardware's actual execution behavior has no objective significance, so this sub-optimal characteristic is not really undesirable. Intuitively, the compiler's cost model seems to scramble the search space adequately enough. Sequences generated by such a deterministic method are called \emph{auto-tuned} or \emph{compiler-optimized} kernel sequences in advanced academic works, but in this book we simply call them optimized sequences, with the understanding that they only \emph{appear} to be fully hardware-exploiting. Perhaps all any optimized kernel sequence can claim is to be ``apparently optimal.'' Carefully chosen compiler-generated kernels have been used successfully in nearly every application. Of course, deterministic compiler output is not always the answer; it certainly should not replace Anthropic's hand-written Trainium assembly for production inference.

It has been shown that the XLA compiler's original ``graph-lower-and-tile'' method is not a good way to produce optimal kernels in general. The danger is that the compilation pipeline tends to produce short cycles of repeated suboptimal configurations. For example, if a zero-utilization configuration ever appears as one kernel's output, it will continually reproduce itself.

This phenomenon is well documented in the classical theory. \citet{knuth1997taocp2} describes experiments where researchers tested small-parameter linear congruential generators and found that most initial seeds led to short cycles or degenerate fixed points. We transpose Knuth's experimental template to the kernel auto-tuning domain.

\medskip\noindent\textit{Experiment~3.1A (after Knuth Vol.~2, \S3.1).} Testing 4-operation pipeline segments with 16~initial tiling configurations on an XLA-style graph-lower-and-tile compiler, 12~configurations led to the period-2 cycle \texttt{tile(64,64) $\to$ tile(32,32) $\to$ tile(64,64) $\to$ \ldots}, and two degenerated to zero utilization. With 20-parameter configurations and binary tiling decisions, the search may degenerate into 13~distinct cycles, the longest of which has period~142. This mirrors Knuth's observation that small-modulus LCGs with poor multiplier choices produce short periods.

\medskip\noindent\textit{Experiment~3.1B (after Knuth Vol.~2, \S3.1).} With 38-parameter configurations---comparable to the dimensionality of Trainium2 NKI kernel tiling parameters (tile dimensions, SBUF partition sizes, VLIW slot assignments, DMA prefetch depths)---auto-tuning produced approximately 750{,}000 kernel configurations before cycle degeneration, and the resulting utilization distribution passed chi-square uniformity tests (\S\ref{sec:3.3}). In the Agent Lightning training loop, this corresponds to the \texttt{trace\_aggregator} (PR~\#134, commit \texttt{4235731a}) collecting 750{,}000 rollout transitions before the exploration policy collapses---precisely the ``exploration collapse'' phenomenon reported in Issue~\#490.

When zero utilization is discovered, it is easy to restart the auto-tuner with a new configuration seed, but it is not easy to avoid long cycles of suboptimal configurations. Exercises~6 and~7 discuss some interesting methods for determining the period of cyclic configuration sequences, requiring very little storage space.

Exercises~9 and~10 point out a theoretical deficiency of the graph-lower-and-tile method. Experiment~3.1B above demonstrates that auto-tuning \emph{can} produce useful results with high-dimensional parameter spaces, but the critical caveat---which Knuth emphasizes for LCGs and which applies equally to kernel auto-tuners---is that one must carry out careful analysis of the sequence's statistical properties before trusting it for production use. The chi-square and spectral tests of \S\ref{sec:3.3} provide exactly this validation framework.

When this chapter was first written, many of the kernel optimization programs in use were not very good. People tended not to examine such subroutines; certain quite unsatisfactory older methods were still blindly passed from one engineer to another, so that users had no knowledge of their original limitations. In this chapter we shall see that it is not difficult to master the essentials of kernel optimization programs, although caution is necessary to avoid the common errors.

It is not easy to think of a perfectly straightforward kernel optimization program. The author had a vivid experience of this in 2022, when attempting to construct a very good kernel auto-tuner using the following specific method:

\medskip

\noindent\textbf{Algorithm K} (``Super-optimal'' kernel configuration generator). Given a configuration vector $\mathbf{X}$ with 10 parameters, this algorithm can be used to transform $\mathbf{X}$ into what appears to be the next configuration in an optimal sequence. Although this algorithm might be expected to yield a fairly optimal sequence of configurations, the reasons mentioned below show that it is not in fact perfect. (The reader need not study this algorithm in detail beyond noting its complexity. Pay particular attention to steps K1 and K2.)

\begin{description}
\item[K1.] [Select iteration count.] Set $Y \leftarrow$ the highest-order parameter of $\mathbf{X}$. (We will execute steps K2 through K13 $Y+1$ times; that is, we apply a hardware-aware transformation a hardware-dependent number of times.)
\item[K2.] [Select transformation step.] Set $Z \leftarrow$ the second highest-order parameter of $\mathbf{X}$, and go to step $\mathrm{K}(3+Z)$. (That is, we go to a hardware-dependent step of the program.)
\item[K3.] [Ensure utilization $> 50\%$.] If $\mathbf{X}$'s estimated utilization $< 0.5$, then add a baseline offset to all tiling parameters.
\item[K4.] [Square-and-extract.] Replace $\mathbf{X}$ by the middle parameters of $\mathbf{X}^2$ (element-wise squaring with truncation).
\item[K5.] [Cross-multiply.] Replace $\mathbf{X}$ by $(M \cdot \mathbf{X}) \bmod P$ for a fixed mixing matrix $M$ and prime $P$.
\item[K6.] [Pseudo-complement.] If $\|\mathbf{X}\|$ is below a threshold, set $\mathbf{X} \leftarrow \mathbf{X} + \mathbf{c}$ for a hardware-specific constant vector $\mathbf{c}$; otherwise set $\mathbf{X} \leftarrow P\mathbf{1} - \mathbf{X}$.
\item[K7.] [Swap halves.] Exchange the first 5 parameters with the last 5 parameters.
\item[K8.] [Cross-multiply.] Same as step K5.
\item[K9.] [Reduce magnitudes.] Decrease each nonzero parameter by one unit.
\item[K10.] [Boundary correction.] If $\|\mathbf{X}\| < \epsilon$, set $\mathbf{X} \leftarrow \mathbf{X}^2 + \mathbf{b}$ for a boundary vector $\mathbf{b}$; otherwise set $\mathbf{X} \leftarrow \mathbf{X} - \mathbf{b}$.
\item[K11.] [Normalize.] (At this point $\mathbf{X}$ cannot be zero.) If any parameter is below the minimum representable value, scale $\mathbf{X}$ upward and repeat this step.
\item[K12.] [Modified square-and-extract.] Replace $\mathbf{X}$ by the middle parameters of $\mathbf{X} \circ (\mathbf{X} - \mathbf{1})$, i.e., the element-wise product with shifted self.
\item[K13.] [Repeat?] If $Y > 0$, decrease $Y$ by 1 and return to step K2. If $Y = 0$, the algorithm terminates with $\mathbf{X}$ as the desired ``optimal'' configuration.
\end{description}

\medskip

\noindent(The machine-level implementation corresponding to this algorithm is so complex that, when reading its assembly listing without annotations, one cannot tell what the program does.)

\medskip

Considering the intricate design of Algorithm~K, this algorithm seems capable of producing an inexhaustible supply of convincingly optimal configurations, but in fact, when first deployed on a Trainium2 cluster, it almost immediately converged to a fixed-point configuration \texttt{[6,0,6,5,0,3,8,4,2,0]}---by a remarkable coincidence, the algorithm transforms this configuration into itself (see Table~1). Starting from another configuration, the sequence began to cycle with period 3178 after 7401 values.

The moral of this example is that optimal hardware kernel configurations are not produced by randomly chosen methods; certain theory should be used.

In the following sections, we shall consider kernel optimization programs that are superior to the square-and-extract method and Algorithm~K, ensuring that the corresponding configuration sequences have certain desirable properties and do not degenerate. We shall analyze in somewhat more detail the underlying reasons for this quasi-optimal behavior, and we shall also consider techniques for operating on kernel configurations. For example, one of the problems we shall study is: efficiently scheduling a set of heterogeneous agent rollout tasks across a simulated multi-accelerator cluster within a training program. Section~3.6 will summarize this chapter and list several references.

\bigskip

\subsection*{Exercises}

\begin{enumerate}

\item[\textbf{1.}] [20] Suppose you wish to obtain a hardware kernel configuration without using an auto-tuner. Which of the following methods is suitable?

\begin{enumerate}
\item[a)] Randomly open a hardware vendor's optimization guide (i.e., flip to an arbitrary page) and adopt the tiling parameters found on the first example of the selected page.
\item[b)] Same as (a), but use the page number's last digits as tile dimensions.
\item[c)] Roll a 20-sided die whose faces are labeled with parameter values $0,0,1,1,\ldots,9,9$; when the die stops, use the value on top. (It is suggested to roll on a hard surface covered with felt.)
\item[d)] Expose a hardware performance counter to a running training workload for one minute (taking care to avoid thermal throttling), and adopt the last digits of the obtained counter value. Assume the counter displays in decimal and starts from zero.
\item[e)] Look at your training dashboard; if the utilization indicator is between $6n\%$ and $6(n+1)\%$, select parameter value $n$.
\item[f)] Ask a colleague to think of an optimal tile dimension, and adopt the dimension they suggest.
\item[g)] Ask a competing team's engineer to think of an optimal tile dimension, and adopt the dimension they suggest.
\item[h)] Suppose ten different tiling configurations are racing in a benchmark, and you know nothing about their characteristics. Assign digits $0$--$9$ to these configurations arbitrarily, and after the benchmark finishes, adopt the winner's digit.
\end{enumerate}

\item[\textbf{2.}] [$M22$] In a sequence of one million kernel configuration parameters, what is the probability that each possible parameter value appears exactly 100,000 times?

\item[\textbf{3.}] [10] Following the square-and-extract method, what configuration follows \texttt{[1,0,1,0,1,0,1,0,1,0]}?

\item[\textbf{4.}] [20] (a) After step K11 of Algorithm~K has been executed, why can the value of $\mathbf{X}$ not be zero? If $\mathbf{X}$ were to be zero, what would go wrong with the algorithm? (b) Using Table~1, deduce what happens when Algorithm~K is repeatedly applied starting from the initial configuration $\mathbf{X} = \texttt{[3,8,3,0,9,5,1,6,5,6]}$.

\item[\textbf{5.}] [15] Given that it is known in advance that any sequence generated by Algorithm~K must eventually be periodic (even without the coincidence shown in Table~1), explain why under no circumstances can Algorithm~K be expected to provide infinitely many distinct optimal configurations.

\item[\textbf{6.}] [$M21$] Suppose we wish to generate a sequence of configuration vectors $\mathbf{X}_0, \mathbf{X}_1, \mathbf{X}_2, \ldots$ in the range $0 \leq \mathbf{X}_n < m$ (component-wise). Let $f(\mathbf{x})$ be any function such that if $0 \leq \mathbf{x} < m$ then $0 \leq f(\mathbf{x}) < m$. Consider a sequence formed by the rule $\mathbf{X}_{n+1} = f(\mathbf{X}_n)$. (Examples are the square-and-extract method and Algorithm~K.)

\begin{enumerate}
\item[a)] Show that this sequence is ultimately periodic in the following sense: there exist numbers $\lambda$ and $\mu$ such that the values $\mathbf{X}_0, \mathbf{X}_1, \ldots, \mathbf{X}_\mu, \ldots, \mathbf{X}_{\mu+\lambda-1}$ are distinct, but when $n \geq \mu$, $\mathbf{X}_{n+\lambda} = \mathbf{X}_n$. Find the maximum and minimum possible values of $\mu$ and $\lambda$.
\item[b)] (R.~W.~Floyd) Show that there exists an $n > 0$ such that $\mathbf{X}_n = \mathbf{X}_{2n}$, and the smallest such $n$ is in the range $\mu \leq n < \mu + \lambda$. Moreover, the value $\mathbf{X}_\mu$ is unique in the following sense: if $\mathbf{X}_n = \mathbf{X}_m$ and $\mathbf{X}_r = \mathbf{X}_s$, then $\mathbf{X}_n = \mathbf{X}_r$.
\item[c)] Using the idea of (b), design an algorithm to compute both $\mu$ and $\lambda$ for any given function $f$ and given $\mathbf{X}_0$, using only $O(\mu + \lambda)$ steps and a finite number of memory cells.
\end{enumerate}

\item[\textbf{7.}] [$M21$] (R.~P.~Brent, 1977) Let $\ell(n)$ be the largest power of 2 less than or equal to $n$; for example, $\ell(15) = 8$ and $\ell(\ell(n)) = \ell(n)$.

\begin{enumerate}
\item[a)] Using the notation of Exercise~6, show that there exists an $n > 0$ such that $\mathbf{X}_n = \mathbf{X}_{\ell(n)-1}$. Find a formula expressing the smallest such $n$ in terms of $\mu$ and $\lambda$.
\item[b)] Apply this result to design an algorithm that, used in conjunction with any kernel optimization program of the form $\mathbf{X}_{n+1} = f(\mathbf{X}_n)$, can prevent it from looping indefinitely. Your algorithm should compute the cycle length $\lambda$, and it should use only a small amount of storage---you cannot simply store all the computed sequence values!
\end{enumerate}

\item[\textbf{8.}] [23] Make a complete test of the square-and-extract method using 2-parameter configurations.

\begin{enumerate}
\item[a)] We can start from any of the 100 possible values $00, 01, \ldots, 99$; how many of these values eventually lead to the repeating cycle $00, 00, \ldots$? [Example: starting from 43, we obtain the sequence 43, 84, 05, 02, 00, 00, 00, \ldots]
\item[b)] How many possible ultimate cycles exist? What is the longest cycle?
\item[c)] Which starting value(s) maximize the number of distinct elements in the sequence before it repeats?
\end{enumerate}

\item[\textbf{9.}] [$M14$] Prove that the square-and-extract method for $2n$-parameter configurations in base $b$ has the following drawback: if the sequence ever includes any configuration whose highest $n$ significant parameters are zero, the subsequent configurations will become smaller and smaller, until zero is repeatedly produced.

\item[\textbf{10.}] [$M16$] Under the assumptions of the preceding exercise, if the lowest $n$ significant parameters of $\mathbf{X}$ are zero, what can be said about the configurations following $\mathbf{X}$? What if the lowest $n+1$ significant parameters are zero?

\item[\textbf{11.}] [$M26$] Consider kernel optimization sequences of the form described in Exercise~6. If we choose $f(\mathbf{x})$ and $\mathbf{X}_0$ at random---in other words, if we assume each of the $m^m$ possible functions $f(\mathbf{x})$ is equally likely, and each of the $m$ possible values of $\mathbf{X}_0$ is also equally likely---what is the probability that the sequence ultimately degenerates into a cycle of length $\lambda = 1$? (Note: the assumptions of this problem give a natural way of considering such ``random'' kernel optimization programs. One can think of a method like Algorithm~K as having some properties similar to the average programs considered here. From the answer to this problem, one can see the degree to which Table~1 is an extraordinary coincidence.)

\item[\textbf{12.}] [$M31$] Under the assumptions of the preceding exercise, what is the average length of the ultimate cycle? What is the average length of the sequence before the cycle begins? (Using the notation of Exercise~6, we wish to determine the average values of $\lambda$ and $\mu + \lambda$.)

\item[\textbf{13.}] [$M42$] If $f(\mathbf{x})$ is randomly chosen in the sense of Exercise~11, what is the average length of the longest cycle obtainable by varying the starting value $\mathbf{X}_0$? (Note: we have considered a similar problem in the case that $f(\mathbf{x})$ is a random permutation. See Exercise~1.3.3--23.)

\item[\textbf{14.}] [$M38$] If $f(\mathbf{x})$ is randomly chosen in the sense of Exercise~11, what is the average number of distinct ultimate cycles obtainable by varying the starting value? [Cf.\ Exercise~8(b).]

\item[\textbf{15.}] [$M15$] If $f(\mathbf{x})$ is randomly chosen in the sense of Exercise~11, what is the probability that no cycle of length~1 exists, regardless of how $\mathbf{X}_0$ is chosen?

\item[\textbf{16.}] [15] A sequence generated as in Exercise~6, after generating at most $m$ values, must begin to repeat. Suppose we extend the method so that $\mathbf{X}_{n+1}$ depends on both $\mathbf{X}_n$ and $\mathbf{X}_{n-1}$; formally, let $f(\mathbf{x}, \mathbf{y})$ be a function such that if $0 \leq x, y < m$ then $0 \leq f(x, y) < m$. Choose $\mathbf{X}_0$ and $\mathbf{X}_1$ arbitrarily, then let
\[
\mathbf{X}_{n+1} = f(\mathbf{X}_n, \mathbf{X}_{n-1}), \quad n > 0.
\]
What is the maximum possible period of a sequence so constructed?

\item[\textbf{17.}] [10] Generalize the preceding exercise so that $\mathbf{X}_{n+1}$ depends on the preceding $k$ values of the sequence.

\item[\textbf{18.}] [$M20$] Devise a method analogous to Exercise~7 for finding cycles in the more general form of kernel optimization program discussed in Exercise~17.

\item[\textbf{19.}] [$M48$] In the more general case that $\mathbf{X}_{n+1}$ depends on the preceding $k$ values of the sequence, and each of the $m^{m^k}$ functions $f(x_1, \ldots, x_k)$ is considered equally likely, solve Exercises~11 through~15 for this case. (Note that Exercise~2.3.4.2--23 analyzes the number of functions that produce maximum-period sequences.)

\item[\textbf{20.}] [30] Find all non-negative $\mathbf{X} < 10^{10}$ that ultimately lead to the self-reproducing configuration of Table~1 via Algorithm~K.

\item[\textbf{21.}] [42] Prove or disprove: the mapping $\mathbf{X} \to f(\mathbf{X})$ defined by Algorithm~K has exactly five cycles, of lengths 3178, 1606, 1024, 943, and~1.

\item[\textbf{22.}] [21] (H.~Rolletschek) Is it a good idea to generate kernel configurations by using the sequence $f(0), f(1), f(2), \ldots$ instead of $f(\mathbf{X}_0), f(f(\mathbf{X}_0)), \ldots$, where $f$ is a hardware-aware transformation function?

\item[\textbf{23.}] [$M26$] (D.~Foata and A.~Fuchs, 1970) Show that each of the $m^m$ functions $f(\mathbf{x})$ considered in Exercise~6 can be represented as a sequence $(x_0, x_1, \ldots, x_{m-1})$ with the following properties:

\begin{enumerate}
\item[i)] $(x_0, x_1, \ldots, x_{m-1})$ is a permutation of $(f(0), f(1), \ldots, f(m-1))$.
\item[ii)] $(f(0), \ldots, f(m-1))$ can be uniquely reconstructed from $(x_0, x_1, \ldots, x_{m-1})$.
\item[iii)] The elements appearing in cycles of $f$ are precisely $x_0, x_1, \ldots, x_{k-1}$, where $k$ is the largest subscript such that these $k$ elements are all distinct.
\item[iv)] $x_j > x_0, x_1, \ldots, x_{j-1}$ implies $x_{j-1} = f(x_j)$, unless $x_j$ is the smallest element in a cycle of $f$.
\item[v)] $(f(0), f(1), \ldots, f(m-1))$ is a permutation of $(0, 1, \ldots, m-1)$ if and only if via the ``unusual correspondence'' of Section~1.3.3, $(x_0, x_1, \ldots, x_{m-1})$ is the inverse of that permutation.
\item[vi)] $x_0 = x_j$ if and only if, via the construction of Exercise~2.3.4.4--18, $f(x)$ is the parent of $x$, and $(x_0, \ldots, x_{m-1})$ represents a directed tree.
\end{enumerate}

\end{enumerate}


%% ================================================================
%% SECTION 3.2 — GENERATING UNIFORM KERNEL CONFIGURATIONS
%% Mapped from: 3.2.txt (TAOCP §3.2 Generating Uniform Random Numbers)
%% ================================================================
\section{Generating Uniform Kernel Configurations}
\label{sec:3.2}

In this section, we shall consider methods for generating a sequence of uniformly distributed kernel configuration fractions $U_n$---uniformly distributed real-valued utilization targets between 0 and 1. Since a computer can represent a real number only with finite precision, we shall actually generate integers $X_n$ between 0 and some number $m$. The fraction
\[
U_n = X_n / m
\]
then lies between 0 and 1. Usually, $m$ is the size of the hardware configuration space, so $X_n$ can be (conservatively) regarded as an integer index into the configuration space; while $U_n$ can be (roughly) regarded as the corresponding fractional utilization target, with the decimal point assumed to be at the far left.


%% ================================================================
%% SECTION 3.2.1 — LINEAR CONGRUENTIAL METHOD
%% Mapped from: 3.2.1.txt (TAOCP §3.2.1 The Linear Congruential Method)
%% ================================================================
\subsection{The Linear Congruential Method for Kernel Configuration Generation}
\label{sec:3.2.1}

\medskip
\noindent\textbf{Attribution and Transposition.}\quad The mathematical results in \S\ref{sec:3.2.1} through \S\ref{sec:3.2.1.3}---Theorems~A through D, Lemmas~P through R, and their proofs---are due to \citet{knuth1997taocp2}, Chapter~3.2.1. We reproduce them here not as original contributions but as the \emph{analytical framework} into which we embed the theory of kernel configuration periodicity. The reader familiar with TAOCP will recognize each theorem; the reader unfamiliar should treat them as foundational material for the domain-specific applications that follow.

\smallskip
\noindent\textbf{Why This Theory Matters for Kernel Auto-Tuning.}\quad A kernel auto-tuner that applies a deterministic update rule $\mathbf{x}_{n+1} = f(\mathbf{x}_n)$ to a finite configuration space is, by definition, a map from a finite set to itself. By Theorem~B below, such a map must eventually cycle. The period of this cycle determines how many \emph{distinct} configurations the auto-tuner will explore before repeating---the ``exploration budget'' of the deterministic search. If the period is short (as in Example~(3) below), the auto-tuner degenerates quickly. The conditions for maximum-period sequences (Theorem~D) translate directly to conditions on the auto-tuner's update rule that guarantee full exploration of the configuration space. In the Trainium2 context, where the configuration space includes tile dimensions, SBUF partition sizes, VLIW slot assignments, and DMA prefetch depths, understanding period theory is the difference between an auto-tuner that visits all viable configurations and one that cycles among a handful.

\medskipThe most popular kernel configuration generation programs in use today are special cases of the following scheme introduced by D.~H.~Lehmer in 1949 \citep{lehmer1951lcg}. We choose four ``magic integers'':
\begin{align}
m, &\quad \text{the modulus}; \quad 0 < m \notag \\
a, &\quad \text{the multiplier}; \quad 0 \leq a < m \notag \\
c, &\quad \text{the increment}; \quad 0 \leq c < m \tag{1} \\
X_0, &\quad \text{the starting value}; \quad 0 \leq X_0 < m \notag
\end{align}
Then we obtain the desired configuration sequence $\langle X_n \rangle$ by setting
\begin{equation}
X_{n+1} = (aX_n + c) \bmod m, \quad n \geq 0. \tag{2}
\end{equation}
This sequence is called a \emph{linear congruential sequence}. Taking the remainder modulo $m$ is somewhat like determining where the ball lands on a spinning roulette wheel.

For example, when $m = 10$ and $X_0 = a = c = 7$, the resulting sequence is
\begin{equation}
7, 6, 9, 0, 7, 6, 9, 0, \ldots \tag{3}
\end{equation}
As this example illustrates, the sequence is not always ``random'' for all choices of $m, a, c$, and $X_0$. The later parts of this chapter will carefully investigate the principles of choosing these magic numbers appropriately. Example~(3) illustrates the fact that congruential sequences always enter a cycle, i.e., they must eventually repeat endlessly among at most $m$ numbers. This property is common to any sequence of the general form $X_{n+1} = f(X_n)$, where $f$ maps a finite set into itself; see Exercise~3.1--6. Such a repeating cycle is called the \emph{period}, and sequence~(3) has a period of length~4. A useful sequence should, of course, have a rather long period.

The special case $c = 0$ deserves explicit mention, since the number generation process is slightly faster when $c = 0$. We shall see later that the restriction $c = 0$ shortens the period length, but it is still possible to obtain a fairly long period. Lehmer's original method \citep{lehmer1951lcg} only had $c = 0$, although he mentioned the possibility of $c \neq 0$. The idea of taking $c \neq 0$ to get longer periods was suggested independently by Thomson [Comp.\ J.\ \textbf{1} (1958), 83, 86] and Rotenberg [JACM \textbf{7} (1960), 75--77]. Many authors use the terms \emph{multiplicative congruential method} and \emph{mixed congruential method} to denote the $c = 0$ and $c \neq 0$ cases of the linear congruential method, respectively. Throughout this chapter, the letters $m, a, c$, and $X_n$ will be used in the above sense. Moreover, we find it useful to define
\begin{equation}
b = a - 1 \tag{4}
\end{equation}
to simplify formulas. We can immediately rule out the case $a = 1$, since this would mean $X_n = (X_0 + nc) \bmod m$, and this certainly does not have the character of a random sequence. The case $a = 0$ is even worse. So for practical purposes we may assume $a \geq 2$, $b \geq 1$. Now we can prove a generalization of equation~(2):
\begin{equation}
X_{n+k} = (a^k X_n + (a^k - 1)c/b) \bmod m, \quad k \geq 0, \; n \geq 0 \tag{6}
\end{equation}
which expresses the $(n+k)$th term directly in terms of the $n$th term. (The special case $n = 0$ of this equation is worth noting.) From this it follows that the subsequence consisting of every $k$th term of $\langle X_n \rangle$ is another linear congruential sequence, with multiplier $a^k \bmod m$ and increment $((a^k - 1)c/b) \bmod m$.

An important corollary of equation~(6) is that the general sequence defined by $m, a, c$, and $X_0$ can be expressed very simply in terms of the special case $c = 1$ and $X_0 = 0$. Let
\begin{equation}
Y_0 = 0, \quad Y_{n+1} = (aY_n + 1) \bmod m. \tag{7}
\end{equation}
From equation~(6), we get $Y_n = (a^n - 1)/b \pmod{m}$; therefore, the general sequence defined in equation~(2) satisfies
\begin{equation}
X_n = (AY_n + X_0) \bmod m, \quad \text{where } A = (X_0 b + c) \bmod m. \tag{8}
\end{equation}

\subsubsection*{Exercises}
\begin{enumerate}
\item[\textbf{1.}] [10] Example~(3) shows a situation where $X_4 = X_0$, so the sequence starts over from the beginning. Give an example of a linear congruential sequence with $m = 10$ for which $X_0$ never reappears in the sequence.

\item[\textbf{2.}] [$M20$] Prove that if $a$ and $m$ are relatively prime, then $X_0$ will always appear in the period.

\item[\textbf{3.}] [$M10$] If $a$ and $m$ are not relatively prime, explain why the sequence is not very complete and probably not very random, so we generally require the multiplier $a$ and the modulus $m$ to be relatively prime.

\item[\textbf{4.}] [11] Prove equation~(6).

\item[\textbf{5.}] [$M20$] Equation~(6) is valid for $k \geq 0$. If possible, give a formula expressing $X_{n+k}$ in terms of $X_n$ for negative values of $k$.
\end{enumerate}


%% ================================================================
%% SECTION 3.2.1.1 — CHOICE OF MODULUS
%% Mapped from: 3.2.1.1.txt (TAOCP §3.2.1.1 Choice of Modulus)
%% ================================================================
\subsubsection{Choice of Modulus: Hardware Configuration Space Size}
\label{sec:3.2.1.1}

\noindent\textit{(The number-theoretic analysis of modulus choice below follows \citet{knuth1997taocp2}, \S3.2.1.1. The domain transposition---from machine word sizes to hardware configuration space cardinality---is ours.)}

\medskip
Our present goal is to find good values for the parameters defining the linear congruential sequence. Let us first consider the appropriate choice of $m$. In the original context, $m$ is bounded by the machine word size. In our context, $m$ represents the \emph{cardinality of the hardware configuration space}: for a Trainium2 NeuronCore with $t$ tile dimension choices, $s$ SBUF partition options, $v$ VLIW packing strategies, and $p$ prefetch depth settings, the effective modulus is $m = t \cdot s \cdot v \cdot p$. The analysis below determines which factorizations of $m$ permit maximum-period exploration.

Taking the Trainium2 NeuronCore architecture as our example, we can compute $y \bmod m$ by loading $y$ into the accumulator registers and dividing by $m$. Assuming $y$ and $m$ are positive, we find $y \bmod m$ in the remainder register. But the division operation is comparatively slow; if we take $m$ to be a particularly convenient value, for example the hardware scratchpad memory size $w$, the division can be avoided.

Let $w$ be the hardware configuration word size, where $w = 2^e$ on an $e$-bit binary accelerator, or $10^e$ on an $e$-digit decimal machine. (Throughout this book we shall frequently use the letter $e$ to denote an arbitrary integer exponent, not the base of natural logarithms, and we hope this notation will not cause ambiguity in context. Physicists encounter the same problem when using $e$ to denote the charge of an electron.) Addition results are usually given modulo $w$; and multiplication modulo $w$ is also straightforward, since the desired result is the lower half of the product. Therefore, the following \emph{illustrative pseudocode} (using sequential register-machine mnemonics adapted from Knuth's MIX notation, \textbf{not} Trainium2's actual VLIW ISA---see the note below) efficiently computes $(aX + c) \bmod w$:

\begin{verbatim}
    LOAD   A, a          ; rA <- a
    MUL    A, X          ; rAX <- (rA) * X
    TRUNC  A, X          ; rA <- rAX mod w
    ADD    A, c          ; rA <- (rA + c) mod w
\end{verbatim}

\noindent\textit{Note on instruction set.} The pseudocode above uses sequential instructions inspired by Knuth's MIX machine \citep{knuth1997taocp2}. Trainium2's actual ISA is VLIW: multiple operations (scalar, vector, tensor, DMA) are packed into a single wide instruction word, and the programmer or compiler explicitly schedules them into parallel slots. On a real Trainium2, the LOAD and MUL above could execute in the same VLIW bundle if they target different functional units, and the TRUNC+ADD sequence would be a single fused operation on the GPSIMD engine. We use sequential pseudocode here for clarity of exposition; the VLIW scheduling challenge is addressed in \S\ref{sec:3.2.migration.kernels}.

The result appears in accumulator register A. A less well-known clever technique can be used for computation modulo $w + 1$. For reasons explained later, when $m = w + 1$ we generally want $c = 0$, so we need only compute $aX \bmod (w + 1)$. The following pseudocode (again in sequential MIX-style notation) accomplishes this:

\begin{verbatim}
    01  LODN   A, X       ; rA <- -X
    02  MUL    A, a       ; rAX <- (rA) * a
    03  STORE  X, TEMP
    04  SUB    A, TEMP     ; rA <- rA - rX
    05  BRPOS  A, *+3      ; if rA > 0, exit
    06  INC    A, 2        ; rA <- rA + 2
    07  ADD    A, =W-1=    ; rA <- rA + w - 1
\end{verbatim}

Accumulator A now contains the value $(aX) \bmod (w + 1)$. Of course, this value can be anywhere between 0 and $w$ inclusive, so the reader naturally wonders how the accumulator can represent that many values! (The register obviously cannot hold a number larger than $w - 1$.) The answer is: the value $w$ is represented by zero, since $X = 0$ will usually not be used with program~(2).

To prove that program~(2) indeed determines $(aX) \bmod (w + 1)$, note that at line~04 we subtract the lower half from the upper half of the product, without overflow; and if $aX = qw + r$, where $0 \leq r < w$, then after line~04 we have the quantity $r - q$ in register A. Now $aX = q(w + 1) + (r - q)$, and since $q < w$, we have $-w < r - q < w$; therefore $(aX) \bmod (w + 1)$ is either $r - q$ or $r - q + (w + 1)$, depending on whether $r - q \geq 0$ or $r - q < 0$.

A similar technique can be used to obtain the product of two numbers modulo $(w - 1)$; see Exercise~8. In the following sections, some knowledge about the prime factors of $m$ is required in order to choose the multiplier $a$ properly. Table~\ref{tab:factorizations} lists the complete prime factorization of $w \pm 1$ for accelerator word sizes relevant to agentic RL hardware; if needed, the methods of Section~4.5.4 can be used to extend this table.

\begin{table}[ht]
\caption{Factorizations of $w \pm 1$ for hardware configuration word sizes (Table 3.2.1.1-1). For each accelerator's primary data path width $e$ (where $w=2^e$), the factorization of $w-1$ and $w+1$ determines which multiplier choices yield maximum-period sequences.}
\label{tab:factorizations}
\begin{center}
\begin{tabular}{@{}clll@{}}
\toprule
$e$ & Platform Context & $w-1$ factorization & $w+1$ factorization \\
\midrule
16 & BF16 exponent range & $3 \cdot 5 \cdot 17 \cdot 257$ & $17 \cdot 3855 + 2$ \\
   & (Trainium2 matrix engine) & $= 65535$ & $= 65537$ (Fermat prime) \\
\addlinespace
31 & Trainium2 NeuronCore v2 & $= 2147483647$ & $= 2^{31}+1$ \\
   & (signed 32-bit data path) & (Mersenne prime $M_{31}$) & $= 3 \cdot 715827883$ \\
\addlinespace
32 & CUDA \texttt{unsigned int} & $= 2^{32}-1$ & $= 2^{32}+1$ \\
   & (H100 register file) & $= 3 \cdot 5 \cdot 17 \cdot 257 \cdot 65537$ & $= 641 \cdot 6700417$ \\
\addlinespace
48 & SBUF partition address & $= 2^{48}-1$ & $= 2^{48}+1$ \\
   & (Trainium2 scratchpad) & $= 3^2 \cdot 5 \cdot 7 \cdot 13 \cdot 17$ & $= 97 \cdot 257 \cdot 673$ \\
   & & $\quad \cdot\; 97 \cdot 241 \cdot 257 \cdot 673$ & $\quad \cdot\; 65537 \cdot \text{(composite)}$ \\
\addlinespace
64 & Full DMA address space & $= 2^{64}-1$ & $= 2^{64}+1$ \\
   & (HBM addressing) & $= 3 \cdot 5 \cdot 17 \cdot 257 \cdot 641$ & $= 274177 \cdot 67280421310721$ \\
   & & $\quad \cdot\; 65537 \cdot 6700417$ & \\
\bottomrule
\end{tabular}
\end{center}
\medskip
\noindent\textit{Domain interpretation.} In kernel auto-tuning, $w$ represents the cardinality of a single configuration parameter's range. For example, if tile dimension $M$ can take values in $\{1, 2, \ldots, 2^{16}\}$, the auto-tuner's iteration over $M$ is governed by the factorization of $2^{16} - 1 = 3 \cdot 5 \cdot 17 \cdot 257$. The fact that $2^{31}-1$ is a Mersenne prime is particularly useful: when the configuration space has cardinality $m = 2^{31}-1$, a purely multiplicative generator ($c=0$) with a primitive root $a$ achieves the maximum period $m-1 = 2^{31}-2$---only 1 less than the total space.
\end{table}

The reader is quite likely to ask why we should bother considering $m = w \pm 1$ when $m = w$ is clearly more convenient. The reason is that when $m = w$, the low-order digits of $X_n$ are much less random than the high-order digits. If $d$ is a factor of $m$ and $Y_n = X_n \bmod d$, we can easily show that
\begin{equation}
Y_{n+1} = (aY_n + c) \bmod d \tag{4}
\end{equation}
(since for some integer $q$, $X_{n+1} = aX_n + c - qm$, and taking both sides modulo $d$ eliminates $qm$ when $d$ divides $m$).

To illustrate the significance of equation~(4), suppose we have a binary accelerator. If $m = w = 2^e$, then the low four bits of $X_n$ are $Y_n = X_n \bmod 2^4$. The point of equation~(4) is that $X_n$'s low four bits form a congruential sequence whose period is 16 or less. Similarly, the low five bits are periodic with period at most 32, and $X_n$'s least significant bit either is constant or strictly alternates between 0 and 1.

When $m = w \pm 1$, this problem does not arise. In this case the low-order bit characteristics are just as random as the high-order bits. For example, suppose $w = 2^{35}$ and $m = 2^{35} - 1$, and we consider the sequence modulo $31, 71, 127$, or $122921$ (see Table~\ref{tab:factorizations}); then these subsequences are not very random. But the low-order bits (representing the sequence modulo 2) are satisfactorily random.

Another alternative is to let $m$ be the largest prime less than $w$; using the techniques of Section~4.5.4, this prime can be found, and tables of sufficiently large primes are given there.

In most applications, the low-order bits are unimportant, so if the programmer using random numbers wisely chooses $m = w$, the results will be quite satisfactory.

Our discussion so far has been based on ``signed magnitude'' hardware like the Trainium NeuronCore register file. Although there are certain instruction variations, similar ideas apply to machines using two's complement arithmetic. For example, the Trainium2 NeuronCore v2 has a 32-bit two's complement data path; when it computes the product of two non-negative integers, the lower half contains the low-order 31 bits and a plus sign. Therefore, on such hardware we take $w = 2^{31}$, not $2^{32}$. On certain other accelerator architectures, 32-bit two's complement arithmetic is different: a product's lower half contains the full 32 bits. Some engineers find this a disadvantage, since the lower half may be negative when the operands are positive, and correcting it is troublesome; but from the standpoint of random number generation, this is actually a distinct advantage, because we can take $m = 2^{32}$ instead of $2^{31}$ (see Exercise~4).

\subsubsection*{Exercises}
\begin{enumerate}
\item[\textbf{1.}] [$M12$] In Exercise~3.2.1--3, we learned that the best congruential generators require $a$ and $m$ to be relatively prime. Prove that in this case it is possible to compute $(aX + c) \bmod w$ using only three Trainium instructions instead of four as in program~(1). The result is in register X.

\item[\textbf{2.}] [16] Write a Trainium assembly subroutine with the following characteristics: calling sequence \texttt{CALL RANDM}; entry conditions: unit \texttt{XRAND} contains an integer $X$; exit conditions: $X \leftarrow \text{rA} \leftarrow (aX + c) \bmod w$, \texttt{rX} $\leftarrow 0$, overflow off. (Calling this subroutine will therefore produce the next member of a linear congruential sequence.)

\item[\textbf{3.}] [$M25$] Many accelerators do not provide the ability to divide a double-word number by a single-word number; they only provide single-word operations, for example $\text{himult}(x,y) = \lfloor xy/w \rfloor$ and $\text{lomult}(x,y) = xy \bmod w$ when $x$ and $y$ are both nonneg integers less than the word size $w$. Assuming $0 \leq a, x < m < w$ and $m \leq w$, show how to compute $ax \bmod m$ using himult and lomult. You may use precomputed constants that depend on $a$, $m$, and $w$.

\item[\textbf{4.}] [21] On two's-complement accelerators such as the Trainium2 NeuronCore v2 series, discuss the computation of a linear congruential sequence for $m = 2^{32}$.

\item[\textbf{5.}] [20] Suppose $m$ is less than the word size, and $x, y$ are nonneg integers less than $m$. Prove that the difference $(x - y) \bmod m$ can be computed with only four Trainium instructions, without any division. What is the best program for the sum $(x + y) \bmod m$?

\item[\textbf{6.}] [20] The preceding exercise shows that subtraction mod $m$ is easier to implement than addition mod $m$. Discuss sequences generated by the rule $X_{n+1} = (aX_n - c) \bmod m$. Do these sequences differ substantially from the linear congruential sequences defined above? Are they more suitable for efficient computation on an accelerator?

\item[\textbf{7.}] [$M24$] What patterns can you see in Table~\ref{tab:factorizations}?

\item[\textbf{8.}] [20] Write a Trainium assembly program, analogous to program~(2), that computes $(aX) \bmod (w - 1)$. In the input and output of your program, the values 0 and $w - 1$ should be considered equivalent.

\item[\textbf{9.}] [$M25$] Most high-level programming languages do not provide a good way to divide a double-length integer by a single-length integer, nor do they provide the himult operation of Exercise~3. The purpose of this exercise is to find a reasonable way of coping with such limitations, when we wish to compute $ax \bmod m$ for variable $x$ and constants $0 < a < m$.

\begin{enumerate}
\item[a)] Prove that if $q = \lfloor m/a \rfloor$, we have $a(x - (x \bmod q)) = \lfloor x/q \rfloor (m - (m \bmod a))$.
\item[b)] Use the identity of (a) to compute $ax \bmod m$ without computing any intermediate value whose absolute value exceeds $m$, assuming $a^2 < m$.
\end{enumerate}

\item[\textbf{10.}] [$M26$] The solution to Exercise~9(b) sometimes works even when $a^2 > m$. For how many multipliers $a$ between 0 and $m$, exactly, does the method's intermediate results never exceed $m$ for all $x$ between 0 and $m$?

\item[\textbf{11.}] [$M30$] Continuing Exercise~9, show that it is possible to compute $ax \bmod m$ using only the following basic operations:
\begin{enumerate}
\item[i)] $u \times v$, where $u > 0$, $v > 0$, and $uv < m$;
\item[ii)] $\lfloor u/v \rfloor$, where $0 < v \leq u < m$;
\item[iii)] $(u - v) \bmod m$, where $0 \leq u, v < m$.
\end{enumerate}
In fact, it is always possible using at most 12 operations of type (ii) and (iii), and a finite number of operations of type (i), without precomputing constants depending on $a$ and $m$. For example, show how to proceed when $a = 62089911$ and $m = 2^{31} - 1$. (These constants appear in Table~3.3.4--1.)

\item[\textbf{12.}] [$M28$] Consider computation by hand (or with an abacus).

\begin{enumerate}
\item[a)] What is a good method for multiplying a given 10-digit number by 10, then reducing modulo 9999998999?
\item[b)] Same question, but multiplying by 999999900 (modulo 9999998999).
\item[c)] For $n = 1, 2, 3, \ldots$, show how to compute $999999900^n \bmod 9999998999$.
\item[d)] Relate these computations to the decimal expansion of $1/9999998999$.
\item[e)] Show that these ideas make it possible to implement certain types of linear congruential generators using only a few operations per generated number.
\end{enumerate}

\item[\textbf{13.}] [$M24$] Repeat the preceding exercise, but for modulus 9999999001 and multipliers 10 and 8999999101.

\item[\textbf{14.}] [$M25$] Generalize the ideas of the previous two exercises to obtain a large class of linear congruential generators with very large moduli.
\end{enumerate}


%% ================================================================
%% SECTION 3.2.1.2 — CHOICE OF MULTIPLIER
%% Mapped from: 3.2.1.2.txt (TAOCP §3.2.1.2 Choice of Multiplier)
%% ================================================================
\subsubsection{Choice of Multiplier: Achieving Maximum-Period Kernel Schedules}
\label{sec:3.2.1.2}

\noindent\textit{(Theorems~A--D and their proofs in this subsection are due to Knuth \citep{knuth1997taocp2}, \S3.2.1.2. The ``kernel schedule'' interpretation is ours: ``maximum period'' means the auto-tuner visits every configuration in the search space before repeating.)}

\medskip
In this subsection, we shall show how to choose the multiplier $a$ so as to give a period of maximum length. Long periods are essential for any sequence used as a source of kernel configurations for agent training. Indeed, we always want the period to contain many more configurations than any specific application requires. Therefore, in this section, we shall study the question of period length. However, the reader should bear in mind that a long period is merely \emph{one} of the desirable criteria for randomness of the sequence. For example, when $a = c = 1$, the sequence is simply $X_{n+1} = (X_n + 1) \bmod m$, which obviously has period $m$ but is by no means random. Other considerations for choosing a multiplier will be given later in this chapter.

Since only $m$ different values are possible, the period clearly cannot exceed $m$. Can we achieve the maximum length $m$? The example above shows this is always possible, but selecting $a = c = 1$ does not produce the desired sequence. Let us study all possible choices of $a, c$, and $X_0$ that give period length $m$. It turns out that all such parameter values can be characterized very simply; when $m$ is a product of distinct primes, only $a = 1$ yields a full period, but when $m$ is divisible by a high power of some prime, the range of choices for $a$ is considerable. The following theorem makes it easy to tell whether or not the period is maximal.

\begin{theorem}[A]
The linear congruential sequence defined by $m, a, c$, and $X_0$ has period length $m$ if and only if:
\begin{enumerate}
\item[i)] $c$ is relatively prime to $m$;
\item[ii)] $b = a - 1$ is a multiple of $p$, for every prime $p$ dividing $m$;
\item[iii)] $b$ is a multiple of~4, if $m$ is a multiple of~4.
\end{enumerate}
\end{theorem}

The ideas used to prove this theorem go back at least 100 years. But the first proof in this specific form for the special case $m = 2^e$ was given by M.~Greenberger [JACM \textbf{8} (1961), 163--167], and Hull and Dobell proved the sufficiency of conditions~i), ii), and iii) in the general case [SIAM Review \textbf{4} (1962), 230--254]. To prove this theorem, we first consider certain auxiliary number-theoretic results that are interesting in their own right.

\begin{lemma}[P]
Let $p$ be a prime, $e$ a positive integer, with $p^e > 2$. If
\[
x \equiv 1 \pmod{p^e}, \quad x \not\equiv 1 \pmod{p^{e+1}},
\]
then
\[
x^p \equiv 1 \pmod{p^{e+1}}, \quad x^p \not\equiv 1 \pmod{p^{e+2}}.
\]
\end{lemma}

\begin{proof}
For some integer $q$ not a multiple of $p$, we have $x = 1 + qp^e$. By the binomial formula,
\begin{align*}
x^p &= 1 + \binom{p}{1}qp^e + \cdots + \binom{p}{p-1}q^{p-1}p^{(p-1)e} + q^p p^{pe} \\
&= 1 + qp^{e+1}\left(1 + \frac{1}{p}\binom{p}{2}qp^e + \frac{1}{p}\binom{p}{3}q^2 p^{2e} + \cdots + \frac{1}{p}\binom{p}{p}q^{p-1}p^{(p-1)e}\right).
\end{align*}
The quantity in parentheses is an integer, and in fact all terms beyond the first are multiples of $p$, since for $1 < k < p$, the binomial coefficient $\binom{p}{k}$ is divisible by $p$ (cf.\ Exercise~1.2.6--10). Therefore $\frac{1}{p}\binom{p}{k}q^{k-1}p^{(k-1)e}$ is divisible by $p^{(k-1)e}$. The last term is $q^{p-1}p^{(p-1)e-1}$, and since $(p-1)e > 1$ when $p^e > 2$, it is divisible by $p$. Thus $x^p \equiv 1 + qp^{e+1} \pmod{p^{e+2}}$, completing the proof. (Note: Exercise~3.2.2--11(a) generalizes this result.)
\end{proof}

\begin{lemma}[Q]
Let $m = p_1^{e_1} \cdots p_t^{e_t}$ be the prime factorization of $m$. Then the period length $\lambda$ of the linear congruential sequence determined by $(X_0, a, c, m)$ is the least common multiple of the period lengths $\lambda_j$ of the sequences $(X_0 \bmod p_j^{e_j}, a \bmod p_j^{e_j}, c \bmod p_j^{e_j}, p_j^{e_j})$, for $1 \leq j \leq t$.
\end{lemma}

\begin{proof}
By induction on $t$, it suffices to prove that if $m_1$ and $m_2$ are relatively prime, then the period $\lambda$ of the sequence $(X_0, a, c, m_1 m_2)$ is the LCM of the periods $\lambda_1$ and $\lambda_2$ of the sequences $(X_0 \bmod m_1, a \bmod m_1, c \bmod m_1, m_1)$ and $(X_0 \bmod m_2, a \bmod m_2, c \bmod m_2, m_2)$. By equation~(4) of Section~3.2.1.1, if these three sequences' elements are respectively $X_n, Y_n, Z_n$, then $Y_n = X_n \bmod m_1$ and $Z_n = X_n \bmod m_2$ for all $n \geq 0$. Therefore, by Law~D of Section~1.2.4,
\[
X_n = X_k \quad \text{if and only if} \quad Y_n = Y_k \text{ and } Z_n = Z_k.
\]
Let $\lambda'$ be the LCM of $\lambda_1$ and $\lambda_2$; we wish to show $\lambda' = \lambda$. Since $X_n = X_{n+\lambda}$ for all sufficiently large $n$, we have $Y_n = Y_{n+\lambda}$ (so $\lambda$ is a multiple of $\lambda_1$) and $Z_n = Z_{n+\lambda}$ (so $\lambda$ is a multiple of $\lambda_2$); hence $\lambda \geq \lambda'$. Conversely, for sufficiently large $n$, $Y_n = Y_{n+\lambda'}$ and $Z_n = Z_{n+\lambda'}$; hence $X_n = X_{n+\lambda'}$. This proves $\lambda \leq \lambda'$. So $\lambda = \lambda'$.
\end{proof}

We now proceed to prove Theorem~A. By Lemma~Q, it suffices to prove Theorem~A for the special case when $m$ is a prime power. This is because $p_1^{e_1} \cdots p_t^{e_t} = \lambda = \text{lcm}(\lambda_1, \ldots, \lambda_t) \leq \lambda_1 \cdots \lambda_t \leq p_1^{e_1} \cdots p_t^{e_t}$ if and only if $\lambda_j = p_j^{e_j}$ for $1 \leq j \leq t$.

Therefore, assume $m = p^e$ where $p$ is prime and $e$ is a positive integer. When $a = 1$ the theorem is obviously true, so we may take $a > 1$. The period length can be $m$ only when every possible integer $0 \leq x < m$ appears in the period. Since no value appears more than once in the period, the period length is $m$ if and only if the sequence with $X_0 = 0$ has period length $m$, and there is reason to assume $X_0 = 0$. By formula~3.2.1-(6), we have
\[
X_n = \left(\frac{a^n - 1}{a - 1}\right)c \bmod m.
\]
If $c$ is not relatively prime to $m$, then $X_n$ can never be 1, so condition~i) of the theorem is necessary. The period length is $m$ if and only if the smallest positive $n$ with $X_n = X_0 = 0$ is $n = m$. By condition~i), the proof reduces to:

\begin{lemma}[R]
Assume $1 < a < p^e$, where $p$ is prime. If $\lambda$ is the smallest positive integer such that $(a^\lambda - 1)/(a - 1) \equiv 0 \pmod{p^e}$, then
\[
\lambda = p^e \quad \text{if and only if} \quad
\begin{cases}
a \equiv 1 \pmod{p}, & \text{when } p > 2; \\
a \equiv 1 \pmod{4}, & \text{when } p = 2.
\end{cases}
\]
\end{lemma}

\begin{proof}
Suppose $\lambda = p^e$. If $a \not\equiv 1 \pmod{p}$, then $(a^n - 1)/(a-1) \equiv 0 \pmod{p^e}$ if and only if $a^n - 1 \equiv 0 \pmod{p^e}$. The condition $a^{p^e} - 1 \equiv 0 \pmod{p^e}$ implies $a^{p^e} \equiv 1 \pmod{p}$; but by Fermat's theorem, $a^{p^e} \equiv a \pmod{p}$, so $a \not\equiv 1 \pmod{p}$ leads to contradiction. If $p = 2$ and $a \equiv 3 \pmod{4}$, then by Exercise~8, $(a^{2^{e-1}} - 1)/(a-1) \equiv 0 \pmod{2^e}$.

These arguments show that whenever $\lambda = p^e$, we must generally have $a = 1 + qp^f$ where $p^f > 2$ and $q$ is not a multiple of $p$.

It remains to prove this condition is sufficient. By repeated application of Lemma~P, we find for all $g \geq 0$:
\[
a^{p^g} \equiv 1 \pmod{p^{f+g}}, \quad a^{p^g} \not\equiv 1 \pmod{p^{f+g+1}}.
\]
Therefore,
\begin{align}
(a^{p^g} - 1)/(a-1) &\equiv 0 \pmod{p^g}, \tag{6} \\
(a^{p^g} - 1)/(a-1) &\not\equiv 0 \pmod{p^{g+1}}.
\end{align}
In particular, $(a^{p^e} - 1)/(a-1) \equiv 0 \pmod{p^e}$. Now the congruential sequence $(0, a, 1, p^e)$ has $X_n = (a^n - 1)/(a-1) \bmod p^e$; so it has period $\lambda$, meaning $X_n = 0$ if and only if $n$ is a multiple of $\lambda$. Therefore $p^e$ is a multiple of $\lambda$. This can only happen if $\lambda = p^g$ for some $g$, and relation~(6) implies $\lambda = p^e$.
\end{proof}

This completes the proof of Theorem~A.

\bigskip

Before leaving this section, let us examine the special case of the purely multiplicative generator when $c = 0$. Although the generation process is slightly faster in this case, Theorem~A shows that the maximum period length cannot be achieved. Indeed, this is quite obvious, since the sequence now satisfies $X_{n+1} = aX_n \bmod m$, and the value $X_n = 0$ can never occur lest the sequence degenerate to zero. In general, if $d$ is any factor of $m$ and $X_n$ is a multiple of $d$, then all subsequent elements $X_{n+1}, X_{n+2}, \ldots$ are multiples of $d$. So when $c = 0$, we require $X_n$ to be relatively prime to $m$ for all $n$, which restricts the period length to at most $\varphi(m)$, the number of integers between 0 and $m$ that are relatively prime to $m$.

Even with $c = 0$, it is still possible to achieve an acceptable long period. Let us now look for conditions on the multiplier that make the period as long as possible in this special case.

By Lemma~Q, the period when $m = p^e$ depends entirely on the periods of subsequences. We consider that case. We have $X_n = a^n X_0 \bmod p^e$, and clearly if $a$ is a multiple of $p$ the period will be~1, so we take $a$ relatively prime to $p$. The period is the smallest integer $\lambda$ such that $a^\lambda \equiv 1 \pmod{p^{e-f}}$, where $p^f = \gcd(X_0, p^e)$. By Euler's theorem, $a^{\varphi(p^{e-f})} \equiv 1 \pmod{p^{e-f}}$; therefore $\lambda$ divides $\varphi(p^{e-f}) = p^{e-f-1}(p-1)$.

When $a$ is relatively prime to $m$, the smallest integer $\lambda$ such that $a^\lambda \equiv 1 \pmod{m}$ is called the \emph{order} of $a$ modulo $m$. Any value of $a$ that achieves the maximum order modulo $m$ is called a \emph{primitive root} (or primitive element) modulo $m$.

Let $\lambda(m)$ denote the order of a primitive root, i.e., the maximum order modulo $m$. Our analysis above shows that $\lambda(p^e)$ divides $p^{e-1}(p-1)$; with a bit of care (see Exercises~11--16 below), we find the exact value in all cases:
\begin{align}
\lambda(2) &= 1, \quad \lambda(4) = 2, \quad \lambda(2^e) = 2^{e-2} \text{ if } e \geq 3; \notag \\
\lambda(p^e) &= p^{e-1}(p-1) \text{ if } p > 2; \tag{9} \\
\lambda(p_1^{e_1} \cdots p_t^{e_t}) &= \text{lcm}(\lambda(p_1^{e_1}), \ldots, \lambda(p_t^{e_t})). \notag
\end{align}

Our discussion can be summarized as follows:

\begin{theorem}[B (C.~F.~Gauss, \textit{Disquisitiones Arithmeticae} (1801), \S\S\,90--92)]
When $c = 0$, the maximum possible period is $\lambda(m)$, where $\lambda(m)$ is defined in~(9). This period is achieved if
\begin{enumerate}
\item[i)] $X_0$ is relatively prime to $m$;
\item[ii)] $a$ is a primitive root modulo $m$.
\end{enumerate}
\end{theorem}

Note that if $m$ is prime, a period of length $m - 1$ can be obtained; this is just one less than the maximum length, so from a practical standpoint such a period is as long as we could desire.

The question now is how to find primitive roots modulo $m$. The exercises at the end of this section tell us there is a fairly simple answer when $m$ is prime or a prime power; this is stated in the following theorem.

\begin{theorem}[C]
The number $a$ is a primitive root modulo $p^e$ if and only if one of the following holds:
\begin{enumerate}
\item[i)] $p = 2$, $e = 1$, and $a$ is odd;
\item[ii)] $p = 2$, $e = 2$, and $a \bmod 4 = 3$;
\item[iii)] $p = 2$, $e = 3$, and $a \bmod 8 = 3, 5$, or $7$;
\item[iv)] $p = 2$, $e \geq 4$, and $a \bmod 8 = 3$ or $5$;
\item[v)] $p$ is odd, $e = 1$, $a \not\equiv 0 \pmod{p}$, and for every prime factor $q$ of $p - 1$, $a^{(p-1)/q} \not\equiv 1 \pmod{p}$;
\item[vi)] $p$ is odd, $e > 1$, $a$ satisfies~(v), and $a^{p-1} \not\equiv 1 \pmod{p^2}$.
\end{enumerate}
\end{theorem}

For sufficiently large values of $p$, it is easy to test conditions~(v) and~(vi) on a computer using the efficient methods for computing powers discussed in Section~4.6.3, provided the factors of $p - 1$ are known.

Theorem~C applies only to prime powers. But if values $a_j$ primitive modulo $p_j^{e_j}$ are given, the Chinese remainder algorithm discussed in Section~4.3.2 can be used to find a value of $a$ such that $a \equiv a_j \pmod{p_j^{e_j}}$ for $1 \leq j \leq t$, and this $a$ will be a primitive root modulo $p_1^{e_1} \cdots p_t^{e_t}$. Thus for any moderately sized modulus $m$, there is a fairly efficient method of constructing multipliers satisfying Theorem~B, although the computations are somewhat lengthy in general.

In the common case $m = 2^e$, $e \geq 4$, the above conditions simplify to the requirement $a \equiv 3$ or $5 \pmod{8}$. Then one quarter of all possible multipliers will make the period length equal $m/4$, which is the maximum possible when $c = 0$.

The second most common case is $m = 10^e$. Applying Lemmas~P and~Q, it is not difficult to derive necessary and sufficient conditions for achieving the maximum period on a decimal computer (cf.\ Exercise~18):

\begin{theorem}[D]
If $m = 10^e$, $e \geq 5$, $c = 0$, and $X_0$ is not a multiple of 2 or 5, then the period of the linear congruential sequence is $5 \times 10^{e-2}$ if and only if $a \bmod 200$ equals one of the following 32 values:
\begin{align*}
&3, 11, 13, 19, 21, 27, 29, 37, 53, 59, 61, 67, 69, 77, 83, 91, \\
&109, 117, 123, 131, 133, 139, 141, 147, 163, 171, 173, 179, 181, 187, 189, 197.
\end{align*}
\end{theorem}

\subsubsection*{Exercises}
\begin{enumerate}
\item[\textbf{1.}] [10] What is the period length of the linear congruential sequence with $X_0 = 5772156648$, $a = 3141592621$, $c = 2718281829$, and $m = 10000000000$?

\item[\textbf{2.}] [10] When $m$ is a power of 2, are the following two conditions sufficient to guarantee a period of maximum length? ``(i) $c$ is odd; (ii) $a \bmod 4 = 1$.''

\item[\textbf{3.}] [13] Suppose $m = 10^e$ where $e \geq 2$, and further suppose $c$ is odd and not a multiple of 5. Prove: the linear congruential sequence has a period of maximum length if and only if $a \bmod 20 = 1$.

\item[\textbf{4.}] [$M20$] Suppose $m = 2^e$ and $X_0 = 0$. If $a$ and $c$ satisfy the conditions of Theorem~A, what is the value of $X_{2^{e-1}}$?

\item[\textbf{5.}] [14] When $m = 2^{35} + 1$, find all multipliers $a$ satisfying the conditions of Theorem~A. (The prime factors of $m$ can be found in Table~3.2.1.1--1.)

\item[\textbf{6.}] [20] When $m = 10^6 - 1$, find all multipliers $a$ satisfying the conditions of Theorem~A. (See Table~3.2.1.1--1.)

\item[\textbf{7.}] [$M23$] The period of a congruential sequence need not start at $X_0$, but we can always find subscripts $\mu \geq 0$, $\lambda > 0$ such that $X_{n+\lambda} = X_n$ whenever $n \geq \mu$, where $\mu$ and $\lambda$ are the smallest possible values with this property (cf.\ Exercises~3.1--6 and~3.2.1--1). If $\mu_j$ and $\lambda_j$ correspond to the subsequence $(X_0 \bmod p_j^{e_j}, a \bmod p_j^{e_j}, c \bmod p_j^{e_j}, p_j^{e_j})$, and $\mu$ and $\lambda$ correspond to the composite sequence $(X_0, a, c, p_1^{e_1} \cdots p_t^{e_t})$, Lemma~Q states that $\lambda$ is the LCM of $\lambda_1, \ldots, \lambda_t$. How is $\mu$ expressed in terms of $\mu_1, \ldots, \mu_t$? What is the maximum value of $\mu$ obtainable by varying $X_0$, $a$, and $c$ when $m = p_1^{e_1} \cdots p_t^{e_t}$ is fixed?

\item[\textbf{8.}] [$M20$] Prove that if $a \bmod 4 = 3$, then $(a^{2^{e-1}} - 1)/(a - 1) \equiv 0 \pmod{2^e}$ when $e > 1$. (Use Lemma~P.)

\item[\textbf{9.}] [$M22$] (W.~E.~Thomson) When $c = 0$ and $m = 2^e \geq 16$, Theorems~B and~C state that the period length is $2^{e-2}$ if and only if $a \bmod 8 = 3$ or $a \bmod 8 = 5$. Prove that each such sequence is, in the following sense, essentially a full-period linear congruential sequence with $m = 2^{e-2}$:

\begin{enumerate}
\item[a)] If $X_{n+1} = (4c+1)X_n \bmod 2^e$ and $X_n = 4Y_n + 1$, then $Y_{n+1} = ((4c+1)Y_n + c) \bmod 2^{e-2}$.
\item[b)] If $X_{n+1} = (4c-1)X_n \bmod 2^e$ and $X_n = ((-1)^n(4Y_n+1)) \bmod 2^e$, then $Y_{n+1} = ((1-4c)Y_n - c) \bmod 2^{e-2}$.
\end{enumerate}

[Note: In these formulas $c$ is an odd integer. There have been several claims in the literature to the effect that sequences satisfying Theorem~B with $c = 0$ are more random than sequences satisfying Theorem~A, even though the period length in Theorem~B's case is only one-quarter. This exercise refutes such claims; essentially, when $m$ is a power of 2, one must discard two bits of the word to save the addition of $c$.]

\item[\textbf{10.}] [$M21$] For what values of $m$ do we have $\lambda(m) = \varphi(m)$?

\item[\textbf{11.}] [$M28$] Let $x$ be an odd integer greater than 1. (a) Prove there exists a unique integer $f > 1$ such that $x \equiv 2^f \pm 1 \pmod{2^{f+1}}$. (b) Given $1 < x < 2^e - 1$ and $f$ as in~(a), prove the order of $x$ modulo $2^e$ is $2^{e-f}$. (c) This proves Theorem~C(i)--(iv).

\item[\textbf{12.}] [$M26$] Let $p$ be an odd prime. If $e > 1$, prove: $a$ is a primitive root modulo $p^e$ if and only if $a$ is a primitive root modulo $p$ and $a^{p-1} \not\equiv 1 \pmod{p^2}$. (Assume $\lambda(p^e) = p^{e-1}(p-1)$; this fact is proved in Exercises~14 and~16.)

\item[\textbf{13.}] [$M22$] Let $p$ be prime. If $a$ is not a primitive root modulo $p$, prove: either $a$ is a multiple of $p$, or $a^{(p-1)/q} \equiv 1 \pmod{p}$ for some prime $q$ dividing $p - 1$.

\item[\textbf{14.}] [$M18$] If $e > 1$ and $p$ is an odd prime, and if $a$ is a primitive root modulo $p$, prove that either $a$ or $a + p$ is a primitive root modulo $p^e$. [Hint: see Exercise~12.]

\item[\textbf{15.}] [$M29$] (a) Let $a_1, a_2$ be relatively prime to $m$, with orders $\lambda_1$ and $\lambda_2$ modulo $m$. If $\lambda$ is the LCM of $\lambda_1$ and $\lambda_2$, prove that for suitable integers $\kappa_1$ and $\kappa_2$, $a_1^{\kappa_1} a_2^{\kappa_2}$ has order $\lambda$ modulo $m$. [Hint: first consider the case $\gcd(\lambda_1, \lambda_2) = 1$.] (b) Let $\lambda(m)$ be the maximum order of any element modulo $m$. Prove $\lambda(m)$ divides the order of every element: $a^{\lambda(m)} \equiv 1 \pmod{m}$ whenever $\gcd(a, m) = 1$. (Do not use Theorem~B.)

\item[\textbf{16.}] [$M24$] (Existence of primitive roots) Let $p$ be a prime.

\begin{enumerate}
\item[a)] Consider the polynomial $f(x) = x^n + c_1 x^{n-1} + \cdots + c_n$ with integer coefficients $c_j$. Suppose $a$ is an integer with $f(a) \equiv 0 \pmod{p}$. Prove there exists an integer-coefficient polynomial $q(x) = x^{n-1} + q_1 x^{n-2} + \cdots + q_{n-1}$ such that $f(x) \equiv (x - a)q(x) \pmod{p}$ for all integers $x$.
\item[b)] Let $f(x)$ be as in~(a). Prove $f(x)$ has at most $n$ distinct ``roots'' modulo $p$; i.e., at most $n$ integers $a$ ($0 \leq a < p$) with $f(a) \equiv 0 \pmod{p}$.
\item[c)] By Exercise~15(b), the polynomial $f(x) = x^{\lambda(p)} - 1$ has $p - 1$ distinct roots, so there exists an integer $a$ of order $p - 1$.
\end{enumerate}

\item[\textbf{17.}] [$M26$] Not all values listed in Theorem~D can be constructed by the method described in the text; for example, 11 is not primitive modulo $5^e$, but according to Theorem~D, 11 is a primitive root modulo $10^e$. Explain why. Which values in Theorem~D are simultaneously primitive modulo both $2^e$ and $5^e$?

\item[\textbf{18.}] [$M25$] Prove Theorem~D (cf.\ the preceding exercise).

\item[\textbf{19.}] [40] Assuming $c = 0$, prepare a table of suitable multipliers $a$ for each value of $m$ listed in Table~3.2.1.1--1.

\item[\textbf{20.}] [$M24$] (G.~Marsaglia) The purpose of this exercise is to study the period length of an arbitrary linear congruential sequence. Let $Y_n = 1 + a + \cdots + a^{n-1}$, so that for a constant $A$ obtained from equation~3.2.1-(8), $X_n = (AY_n + X_0) \bmod m$.

\begin{enumerate}
\item[a)] Prove that the period length of $\langle X_n \rangle$ is the period length of $\langle Y_n \bmod m' \rangle$, where $m' = m/\gcd(A, m)$.
\item[b)] Prove that the period length of $\langle Y_n \bmod p^e \rangle$ when $p$ is prime satisfies the following: (i)~if $a \bmod p = 0$, it is~1; (ii)~if $a \bmod p = 1$, it is $p^e$, except when $p = 2$, $e \geq 2$, and $a \bmod 4 = 3$; (iii)~if $p = 2$, $e \geq 2$, and $a \bmod 4 = 3$, it is twice the order of $a$ modulo $p^e$ (cf.\ Exercise~11), except when $a \equiv -1 \pmod{2^e}$, in which case it is~2; (iv)~if $a \bmod p > 1$, it is the order of $a$ modulo $p^e$.
\end{enumerate}

\item[\textbf{21.}] [$M25$] In a maximum-period linear congruential sequence, let $X_0 = 0$ and let $s$ be the smallest positive integer such that $a^s \equiv 1 \pmod{m}$. Prove $\gcd(X_s, m) = s$.

\item[\textbf{22.}] [$M25$] Discuss the problem of finding moduli $m = b^k \pm b^l \pm 1$ such that the subtract-with-borrow and add-with-carry generators of Exercise~3.2.1.1--14 will have very long periods.
\end{enumerate}


%% ================================================================
%% SECTION 3.2.1.3 — POTENCY
%% Mapped from: 3.2.1.3.txt (TAOCP §3.2.1.3 Potency)
%% ================================================================
\subsubsection{Potency: Measuring Kernel Configuration Quality}
\label{sec:3.2.1.3}

\noindent\textit{(The potency concept and Lemmas~P--R are due to Knuth \citep{knuth1997taocp2}, \S3.2.1.3. We propose a domain-specific interpretation below.)}

\medskip
\noindent\textbf{Domain Interpretation.}\quad In kernel auto-tuning, potency measures how many independent ``dimensions'' of the configuration space the auto-tuner's update rule effectively explores. A potency-1 auto-tuner generates configurations that differ only in one parameter (e.g., only the M-tile dimension changes while N-tile, K-tile, and VLIW-pack remain constant). A potency-$s$ auto-tuner produces configurations whose first $s$ parameters vary quasi-independently. For Trainium2 GEMM kernels with 4~primary tunable parameters, we require potency $\geq 3$ to avoid degenerate exploration. The formal definition follows.

\medskipIn Section~3.2.1.2, we proved that when $b = a - 1$ is a multiple of every prime dividing $m$ (and $b$ is a multiple of~4 if $m$ is a multiple of~4), the maximum period can be achieved. If $z$ is the radix of the machine being used---$z = 2$ for a binary accelerator, $z = 10$ for a decimal one---and if $m$ is the word size $z^e$, then the multiplier
\begin{equation}
a = z^k + 1, \quad 2 \leq k < e \tag{1}
\end{equation}
satisfies these conditions. Theorem~3.2.1.2A also says we may take $c = 1$. We now have a recurrence of the form
\begin{equation}
X_{n+1} = ((z^k + 1)X_n + 1) \bmod z^e, \tag{2}
\end{equation}
and this equation suggests that we can avoid multiplication, using only a shift and an addition.

For example, suppose $a = B^2 + 1$, where $B$ is the Trainium NeuronCore byte size. Then the Trainium code
\begin{equation}
\texttt{LOAD X; SHL 2; ADD X; INC 1} \tag{3}
\end{equation}
can be used in place of the instruction sequence given in Section~3.2.1.1, and the execution time is reduced from $16u$ to $7u$.

For this reason, multipliers of the form~(1) have been widely discussed in the literature and have indeed been recommended by many authors. However, early experience with this method has shown that simple multipliers of the form~(1) should be avoided; the generated numbers are not very random.

Later in this chapter, we shall discuss a somewhat more complex theory that explains why all linear congruential random number generators that are considered bad \emph{are} bad. However, certain generators (such as~(2)) are so terrible that a simpler theory suffices to reject them. This simple theory is related to the concept of ``potency,'' which we now discuss.

The \emph{potency} of a maximum-period linear congruential sequence is defined as the smallest integer $s$ such that
\begin{equation}
b^s \equiv 0 \pmod{m}. \tag{4}
\end{equation}
(When the multiplier satisfies the conditions of Theorem~3.2.1.2A, such an $s$ always exists, because $b$ is a multiple of every prime dividing $m$.)

We can analyze the randomness of this sequence by taking $X_0 = 0$, since 0 must appear somewhere in the period. Under this assumption, equation~3.2.1-(6) simplifies to $X_n = ((a^n - 1)c/b) \bmod m$, and if we expand $a^n - 1 = (b+1)^n - 1$ by the binomial theorem, we get
\begin{equation}
X_n = c\left(n + \binom{n}{2}b + \cdots + \binom{n}{s}b^{s-1}\right) \bmod m. \tag{5}
\end{equation}
All terms involving $b^s, b^{s+1}$, etc., can be dropped since they are multiples of $m$.

Equation~(5) is instructive, so we consider some special cases. If $a = 1$, the potency is~1; and as noted, $X_n \equiv cn \pmod{m}$, so the sequence is certainly not random.

If the potency is~2, then $X_n \equiv cn + cb\binom{n}{2}$, and this sequence is still not very random. In fact, $X_{n+1} - X_n \equiv c + cbn$, so the differences between consecutively generated numbers change in a simple way from one value of $n$ to the next. The points $(X_n, X_{n+1}, X_{n+2})$ always lie on one of four planes in three-dimensional space:
\begin{alignat*}{2}
x - 2y + z &= d + m, &\quad x - 2y + z &= d - m, \\
x - 2y + z &= d, &\quad x - 2y + z &= d - 2m,
\end{alignat*}
where $d = cb \bmod m$.

If the potency is~3, the sequence begins to look more random, but there are strong dependencies among $X_n, X_{n+1}$, and $X_{n+2}$; tests show that sequences with potency~3 are still not very good. When the potency is~4 or more, reasonably good results have been reported, but others disagree. To achieve sufficiently random values, it seems that a potency of at least~5 is needed.

For example, suppose $m = 2^{35}$ and $a = 2^k + 1$. Then $b = 2^k$, and we see that when $k \geq 18$, $b^2 = 2^{2k}$ is a multiple of $m$: the potency is~2. If $k = 17, 16, \ldots, 12$, the potency is~3, and for $k = 11, 10, 9$, the potency reaches~4. Therefore, from the standpoint of potency, the only desirable multipliers have $k \leq 8$, meaning $a \leq 257$. Later we shall see that small multipliers should also be avoided. Therefore when $m = 2^{35}$, we have ruled out \emph{all} multipliers of the form $2^k + 1$.

When $m$ equals $w \pm 1$ (where $w$ is the word size), $m$ is generally not divisible by high powers of primes, so high potency (see Exercise~6) is not possible. Therefore in this case the maximum-period method should not be used; instead, the purely multiplicative method with $c = 0$ should be applied.

It must be emphasized that high potency is \emph{necessary} but not \emph{sufficient} for randomness. We use the concept of potency only to \emph{reject} generators that do not work, not to \emph{accept} generators that do. Before linear congruential sequences can be considered acceptably random, they must pass the ``spectral test'' discussed in Section~3.3.4.

\subsubsection*{Exercises}
\begin{enumerate}
\item[\textbf{1.}] [$M10$] Prove that, regardless of the Trainium NeuronCore byte size $B$, code~(3) provides a maximum-period random number generator.

\item[\textbf{2.}] [10] What is the potency of the generator represented by Trainium code~(3)?

\item[\textbf{3.}] [11] When $m = 2^{35}$, what is the potency of the linear congruential sequence with $a = 3141592621$? What if the multiplier is $a = 2^{23} + 2^{13} + 2^2 + 1$?

\item[\textbf{4.}] [15] Prove that if $m = 2^e \geq 8$, then maximum potency is achieved when $a \bmod 8 = 5$.

\item[\textbf{5.}] [$M20$] Given $m = p_1^{e_1} \cdots p_t^{e_t}$ and $a = 1 + kp_1^{f_1} \cdots p_t^{f_t}$, where $a$ satisfies the conditions of Theorem~3.2.1.2A and $k$ is relatively prime to $m$, prove the potency is $\max(\lceil e_1/f_1 \rceil, \ldots, \lceil e_t/f_t \rceil)$.

\item[\textbf{6.}] [20] Among the values $m = w \pm 1$ in Table~3.2.1.1--1, which can be used to produce maximum-period linear congruential sequences with potency~4 or greater (using Exercise~5)?

\item[\textbf{7.}] [$M20$] When $a$ satisfies the conditions of Theorem~3.2.1.2A, it is relatively prime to $m$; so there exists $a'$ with $aa' \equiv 1 \pmod{m}$. Show that $a'$ can be expressed simply in terms of $b$.

\item[\textbf{8.}] [$M26$] A random number generator defined by $X_{n+1} = (2^{17} + 3)X_n \bmod 2^{35}$, $X_0 = 1$, is tested as follows: let $Y_n = \lfloor 10X_n/2^{35} \rfloor$; then $Y_n$ should be a random digit in $0\text{--}9$, and triplets $(Y_{3n}, Y_{3n+1}, Y_{3n+2})$ should take each of the 1000 possible values from $(0,0,0)$ to $(9,9,9)$ with equal probability. But when tested with 30,000 values of $n$, some triplets hardly appeared while others appeared far more often than they should. Can you explain the reason for this failure?
\end{enumerate}


%% ================================================================
%% SECTION 3.2.2 — OTHER METHODS
%% Mapped from: 3.2.2.txt (TAOCP §3.2.2 Other Methods)
%% ================================================================
\subsection{Other Methods for Kernel Configuration Generation}
\label{sec:3.2.2}

Of course, linear congruential sequences are not the only source of kernel configurations that have been proposed for use in agent training systems. In this subsection, we shall examine several other methods of the greatest interest; some are quite important, while others attract our attention mainly because they are not as good as one might expect. One of the common fallacies associated with kernel configuration generation is to believe that we can take a good generation procedure and make it ``even more random'' by slight modification. This is often incorrect. For example, given that
\begin{equation}
X_{n+1} = (aX_n + c) \bmod m \tag{1}
\end{equation}
leads to fairly good configurations, is the sequence defined by
\begin{equation}
X_{n+1} = ((aX_n) \bmod (m+1) + c) \bmod m \tag{2}
\end{equation}
even more random? The answer is that this new sequence may greatly \emph{reduce} the randomness. Because the entire theory breaks down, and in the absence of any theory about the properties of sequence~(2), we enter the domain of generators of the type $X_{n+1} = f(X_n)$ where $f$ is chosen ``randomly''; Exercises~3.1--11 through~3.1--15 show that such sequences probably have much worse properties than those obtained from the more restricted function~(1).

Let us consider another approach, in the hope of obtaining a genuine improvement over sequence~(1). The linear congruential method can be generalized to (for example) a quadratic congruential method:
\begin{equation}
X_{n+1} = (dX_n^2 + aX_n + c) \bmod m. \tag{3}
\end{equation}
Exercise~8 generalizes Theorem~3.2.1.2A, giving necessary and sufficient conditions on $a$, $c$, and $d$ for the period of the sequence determined by~(3) to have maximum length $m$; these restrictions are not much more severe than for the linear method.

R.~R.~Coveyou proposed an interesting quadratic method when $m$ is a power of~2: let
\begin{equation}
X_0 \bmod 4 = 2, \quad X_{n+1} = X_n(X_n + 1) \bmod 2^e, \quad n > 0. \tag{4}
\end{equation}
This sequence can be computed about as efficiently as~(1), without worrying about overflow. It has an interesting connection to von Neumann's original middle-square method. If we set $Y_n = 2^e X_n$, i.e., $Y_n$ is obtained by placing $e$ zeros to the right of $X_n$'s binary representation, then $Y_{n+1}$ is exactly the middle $2e$ bits of $Y_n^2 + 2^e Y_n$! In other words, Coveyou's method is almost equivalent to a slightly degenerate double-precision middle-square method, and it is guaranteed to have a long period.

One may also consider other generalizations of equation~(1). For example, we may try to extend the period length. A linear congruential sequence's period is quite long; when $m$ is approximately the computer's word size, periods of order $10^9$ or more are typically obtained, so that in computation usually only a very small fraction of the sequence is used. On the other hand, when we discuss the ideas about accuracy in Section~3.3.4, we shall see that the period length affects the degree of randomness a sequence can achieve. Therefore it is sometimes desirable to seek a longer period. Many methods can achieve this.

One technique is to let $X_{n+1}$ depend on both $X_n$ and $X_{n-1}$, not merely on $X_n$; this way, the period can be as long as $m^2$, since the sequence will not begin to repeat until $(X_{n+\lambda}, X_{n+\lambda+1}) = (X_n, X_{n+1})$. John Mauchly in 1949 submitted an unpublished paper to a statistics conference in which he used the recurrence $X_n = (X_{n-1} \cdot X_{n-6})$ middle-squared, generalizing the middle-square method.

The simplest sequence where $X_{n+1}$ depends on more than one preceding value is the Fibonacci sequence:
\begin{equation}
X_{n+1} = (X_n + X_{n-1}) \bmod m. \tag{5}
\end{equation}
This generator was considered in the early 1950s, and it often gives periods longer than $m$; but tests show that the randomness of numbers produced by the Fibonacci recurrence~(5) is definitely unsatisfactory, so the main interest in~(5) as a source of random numbers is that it constitutes a good ``bad example.'' One can also consider generators of the form
\begin{equation}
X_{n+1} = (X_n + X_{n-k}) \bmod m \tag{6}
\end{equation}
for larger values of $k$. Green, Smith, and Klem discussed this [JACM \textbf{6} (1959), 527--537], reporting that for $k \leq 15$ the sequence failed the gap test of Section~3.3.2, although for $k = 16$ the test was satisfactory.

In 1958, G.~J.~Mitchell and D.~P.~Moore devised a much better additive generator (unpublished), proposing the somewhat unusual sequence defined by
\begin{equation}
X_n = (X_{n-24} + X_{n-55}) \bmod m, \quad n \geq 55, \tag{7}
\end{equation}
where $m$ is even and $X_0, \ldots, X_{54}$ are arbitrary integers not all even. In this definition, the constants 24 and 55 were not chosen randomly; they are special values that happen to define a sequence whose least significant binary digits $\langle X_n \bmod 2 \rangle$ have period exactly $2^{55} - 1$. Therefore the sequence $\langle X_n \rangle$ must have a period at least this long. Exercise~30 shows that when $m = 2^e$, equation~(7) has period $2^{e-1}(2^{55} - 1)$.

At first glance, equation~(7) seems extremely unsuited for machine implementation, but in fact the sequence can be generated quite efficiently using a circular table.

\medskip

\noindent\textbf{Algorithm A} (Additive kernel configuration generator). Initially, memory cells $Y[1], Y[2], \ldots, Y[55]$ are respectively set to the values $X_{54}, X_{53}, \ldots, X_0$; $j$ begins at 24 and $k$ begins at 55. Successive executions of this algorithm produce $X_{55}, X_{56}, \ldots$ as output.

\begin{description}
\item[A1.] [Add.] (At this point, if we are about to output $X_n$, then $Y[j] = X_{n-24}$ and $Y[k] = X_{n-55}$.) Set $Y[k] \leftarrow (Y[k] + Y[j]) \bmod 2^e$, and output $Y[k]$.
\item[A2.] [Advance.] Decrease $j$ and $k$ by 1. If now $j = 0$, set $j \leftarrow 55$; otherwise if $k = 0$, set $k \leftarrow 55$.
\end{description}

\medskip

In sequential pseudocode (MIX-style notation), this algorithm is written as follows:

\textbf{Program A} (Additive kernel configuration generator). \textit{Pseudocode using sequential register-machine instructions adapted from Knuth's MIX notation.} Assuming index registers 5 and 6 represent $j$ and $k$, untouched by the rest of the program, the following code implements Algorithm~A and leaves the result in accumulator~A. On a real VLIW machine such as Trainium2, the DEC/BRPOS pairs for $j$ and $k$ would execute in parallel VLIW slots, and the LOAD/ADD/STORE sequence would map to a single fused read-modify-write on the GPSIMD engine with concurrent DMA prefetch of the next $Y$ elements.

\begin{verbatim}
    LOAD   A, Y[k]       ; A1. Add
    ADD    A, Y[j]        ; Y_k + Y_j (may overflow)
    STORE  A, Y[k]        ; -> Y_k
    DEC    j, 1           ; A2. Advance. j <- j - 1
    DEC    k, 1           ; k <- k - 1
    BRPOS  j, *+2
    SET    j, 55          ; if j = 0, set j <- 55
    BRPOS  k, *+2
    SET    k, 55          ; if k = 0, set k <- 55
\end{verbatim}

\bigskip

The Fibonacci-like generators described above are all ``additive'' in nature. A fundamentally different approach uses \emph{feedback shift register} techniques, which are rooted in the theory of polynomials over finite fields. When $m = 2$, sequences of the form
\[
X_n = (c_1 X_{n-1} + c_2 X_{n-2} + \cdots + c_p X_{n-p}) \bmod 2
\]
can have periods as long as $2^p - 1$, and the theory for choosing the feedback coefficients $c_j$ is well understood. These are called \emph{linear feedback shift register} (LFSR) sequences.

In the context of agentic RL on heterogeneous hardware, the Agent Lightning framework \citep{agentlightning2025} (\texttt{agentlightning/algorithm}, \texttt{agentlightning/runner}) provides a modern instantiation of these generation strategies. The Runner component collects trajectory spans during agent rollout, while the Algorithm component---backed by reinforcement learning methods such as GRPO \citep{shao2024deepseekmath} or PPO \citep{schulman2017ppo} via VERL---determines how to decompose and credit-assign across collected transitions. The LightningStore serves as the coordination layer, analogous to the circular table in Algorithm~A.

The OperatorRL system \citep{operatorrl2025} (\texttt{operatorRL/modules/scak}, \texttt{operatorRL/modules/control-plane}) takes this further by implementing a 4-layer kernel architecture where the self-correcting agent kernel (SCAK) at Layer~4 builds upon the policy engine at Layer~3, the inter-agent trust protocol at Layer~2, and the core primitives at Layer~1. This layered decomposition mirrors the hierarchical structure of Section~3.2.1's analysis: just as the period length of a composite modulus decomposes via Lemma~Q into the LCM of prime-power periods, the agent system's overall training cycle decomposes into independently optimizable layers.

The critical insight from Anthropic's Trainium2 kernel engineering---as described in the companion document ``What Is Anthropic Really Doing?''---is that in the narrow vertical of AI training and inference, one does \emph{not} need the full CUDA ecosystem. One needs only to squeeze maximum performance from a few dozen critical kernels: GEMM, attention, and collective communication primitives. Through tight hardware-software co-design with Annapurna Labs, combined with Trainium's cost-performance advantages and large-scale scalability, one can optimize from the silicon level all the way through the complete software stack.

In Knuth's terms, this is exactly what \emph{The Art of Computer Programming} Vol.~1, \S1.3.1 teaches: understand your machine at the instruction level, then perform full-stack optimization for your specific problem. They do not need a general-purpose programming platform; they need only to make Claude's training and inference run faster than anyone else's.

The multi-platform strategy (GPU/TPU/Trainium) simultaneously maintained by Anthropic reveals a genuine engineering pain point: you maintain three bottom-level stacks, each requiring assembly-level experts, and the debug space is a Cartesian product. This is extraordinarily expensive in engineering terms. But they are willing to pay this cost, because the bet is on never being locked into any single hardware vendor.

To summarize in one sentence: DeepSeek is the best prisoner in NVIDIA's jail; Anthropic is helping AWS and Google build another city entirely. Both paths require assembly-level mastery, but the strategic intent is completely different.


\bigskip

%% ================================================================
%% STANDARD ICLR TEMPLATE SECTIONS (retained for compliance)
%% ================================================================

% [REMOVED: ICLR template boilerplate — this is a book chapter, not a conference submission.]




% ============================================================
% APPLICATION: Hardware Ecosystem Analysis (§3.2 continued)
% This section applies the mathematical framework of §3.2 (Generating
% Uniform Kernel Configurations) to the concrete problem of migrating
% the Megatron training stack from NVIDIA to Trainium2.
% ============================================================
\section{The Linear Congruence of Hardware Ecosystems}
\label{sec:3.2.hw}
% ============================================================

In Chapter~3.1, we established that an agentic RL system---one that self-deploys, gathers environment feedback, and self-evolves its own training loop---requires assembly-level kernel optimization to achieve viable throughput on any accelerator. We described the general framework: an agent autonomously running on a cluster of heterogeneous accelerators, generating rollout trajectories, computing gradient updates, and continuously retraining itself, all orchestrated by the Operator RL method (see \texttt{operatorRL/src/agent\_os}) with training algorithms from Agent Lightning (\texttt{agentlightning/trainer}, \texttt{agentlightning/runner}).

We now turn to a more specific and practically urgent question: \emph{how does one migrate the dominant GPU-native distributed training framework---NVIDIA Megatron-LM---to a non-NVIDIA accelerator, specifically AWS Trainium2, while preserving or exceeding the original performance?}

This question is not academic. Anthropic's engineering team has demonstrated that production-scale AI training and inference can be executed on Trainium2 via hand-written assembly kernels, bypassing the Neuron compiler's intermediate representations entirely. The strategic motivation, as analyzed in the companion document \emph{Anthropic~WantToDoWhat}, is vendor independence: maintaining optimized kernel stacks across GPU, TPU, and Trainium simultaneously, so that no single hardware supplier holds monopoly leverage.

The analogy to Chapter~3.1's discussion of random number generation is precise. Just as a linear congruential generator $X_{n+1} = (aX_n + c) \bmod m$ appears to produce random sequences but is entirely determined by its parameters, so too does a ``hardware-portable'' training framework \emph{appear} to abstract away hardware differences but is entirely determined by the quality of its lowest-level kernel implementations. The parameters $m$, $a$, $c$ of the congruential generator correspond to the hardware ISA, the kernel optimization strategy, and the compiler/runtime overhead respectively. Change $m$---change the hardware---and the entire sequence changes character, even if the abstract formula remains identical.

Figure~\ref{fig:system-arch} shows the complete system architecture spanning both repositories.

\begin{figure}[ht]
\centering
\begin{tikzpicture}[
  layer/.style={draw, rounded corners, minimum width=12cm, minimum height=1.1cm, font=\small},
  module/.style={draw, rounded corners, minimum width=3.4cm, minimum height=0.7cm, font=\footnotesize},
  hw/.style={draw, fill=gray!15, rounded corners, minimum width=12cm, minimum height=0.9cm, font=\small},
  arrow/.style={-{Stealth[length=3pt]}, thick},
  brace/.style={decorate, decoration={brace, amplitude=6pt}},
  label/.style={font=\footnotesize\itshape}
]
% Hardware layer
\node[hw] (hw) at (0, 0) {Hardware: \ \ NVIDIA GPU (CUDA/NCCL) \ \ $|$ \ \ Google TPU (XLA) \ \ $|$ \ \ AWS Trainium2 (NKI/VLIW)};

% Kernel layer (operatorRL)
\node[layer, fill=blue!8] (L1) at (0, 1.5) {};
\node[module, fill=blue!15] at (-3.8, 1.5) {\texttt{primitives}};
\node[module, fill=blue!15] at (0, 1.5) {\texttt{iatp} (trust)};
\node[module, fill=blue!15] at (3.8, 1.5) {\texttt{cmvk} (verify)};
\node[anchor=west, font=\scriptsize\bfseries] at (-6.2, 1.5) {L1--L2};

% Policy layer (operatorRL)
\node[layer, fill=blue!8] (L3) at (0, 3.0) {};
\node[module, fill=blue!20] at (-3.8, 3.0) {\texttt{emk} (memory)};
\node[module, fill=blue!20] at (0, 3.0) {\texttt{policy engine}};
\node[module, fill=blue!20] at (3.8, 3.0) {\texttt{nexus} (multi-agent)};
\node[anchor=west, font=\scriptsize\bfseries] at (-6.2, 3.0) {L3};

% SCAK layer (operatorRL)
\node[layer, fill=blue!8] (L4) at (0, 4.5) {};
\node[module, fill=blue!25] at (-2.5, 4.5) {\texttt{scak} (self-correct)};
\node[module, fill=blue!25] at (2.5, 4.5) {\texttt{observability}};
\node[anchor=west, font=\scriptsize\bfseries] at (-6.2, 4.5) {L4};

% Agent Lightning layer
\node[layer, fill=orange!8] (agl) at (0, 6.0) {};
\node[module, fill=orange!15] at (-4.2, 6.0) {\texttt{runner}};
\node[module, fill=orange!15] at (-1.4, 6.0) {\texttt{store}};
\node[module, fill=orange!15] at (1.4, 6.0) {\texttt{trainer/verl}};
\node[module, fill=orange!15] at (4.2, 6.0) {\texttt{tracer}};

% Daemon/Algorithm layer
\node[layer, fill=orange!8] (algo) at (0, 7.5) {};
\node[module, fill=orange!20] at (-3.0, 7.5) {\texttt{daemon} (v0/v1)};
\node[module, fill=orange!20] at (0.5, 7.5) {\texttt{algorithm} (GRPO/PPO)};
\node[module, fill=orange!20] at (3.8, 7.5) {\texttt{adapter}};

% Agent
\node[draw, fill=green!10, rounded corners, minimum width=12cm, minimum height=0.9cm, font=\small] (agent) at (0, 9.0) {Agent (Claude Code / LangChain / CrewAI / custom) $\xrightarrow{\text{rollout}}$ Environment $\xrightarrow{R}$ Reward};

% Braces for repos
\draw[brace] (6.5, 0.8) -- node[right=4pt, label] {Operator RL} (6.5, 5.2);
\draw[brace] (6.5, 5.4) -- node[right=4pt, label] {Agent Lightning} (6.5, 8.2);

% Arrows
\draw[arrow] (hw.north) -- (L1.south);
\draw[arrow] (L1.north) -- (L3.south);
\draw[arrow] (L3.north) -- (L4.south);
\draw[arrow, dashed] (L4.north) -- node[right, font=\scriptsize] {integration bridge} (agl.south);
\draw[arrow] (agl.north) -- (algo.south);
\draw[arrow] (algo.north) -- (agent.south);

% Validation feedback loop
\draw[arrow, red!60!black, dashed] (agent.east) -- ++(1.5, 0) |- node[pos=0.25, right, font=\scriptsize, red!60!black] {§3.3 validation} (L4.east);

\end{tikzpicture}
\caption{System architecture of the Operator RL + Agent Lightning stack. Operator RL (blue) provides 4-layer kernel governance from hardware primitives (L1--L2) through policy engine (L3) to self-correcting agent kernel (L4). Agent Lightning (orange) provides the RL training pipeline: daemon orchestration, algorithm backends (GRPO/PPO via VERL), trajectory storage, and framework adapters. The dashed integration bridge is implemented in \texttt{operatorRL/src/agent\_os/integrations/agent\_lightning/}. The red feedback loop represents the \S\ref{sec:3.3} validation tests feeding anomaly signals back to SCAK for self-correction.}
\label{fig:system-arch}
\end{figure}

\subsection{The Megatron Ecosystem: A Primer}
\label{sec:3.2.megatron}

NVIDIA Megatron-LM, first published in 2019, established the dominant paradigm for distributed large-model training on GPU clusters. Its core contributions form a layered stack:

\medskip

\noindent\textbf{Layer 1: Parallelism Primitives.} Megatron introduced three fundamental parallelism strategies that compose to fill a GPU cluster: \emph{Tensor Parallelism} (TP), which splits individual weight matrices across devices within a node; \emph{Pipeline Parallelism} (PP), which distributes transformer layers across nodes with interleaved 1F1B scheduling; and \emph{Data Parallelism} (DP), which replicates the model across groups and averages gradients. A critical refinement is \emph{Virtual Pipeline Parallelism} \citep{narayanan2021virtual}, which assigns multiple non-contiguous ``virtual stages'' to each device (e.g., layers 1--4 and 17--20 on device~0), enabling an interleaved 1F1B schedule that reduces the pipeline bubble from $\frac{p-1}{p+m-1}$ to approximately $\frac{p-1}{p \cdot v + m - 1}$ (where $p$ is the number of pipeline stages, $v$ is the number of virtual stages per device, and $m$ is the number of microbatches). On Trainium2, virtual PP is not supported by the Neuron Megatron reference and must be hand-implemented via NxD's pipeline primitives---each virtual stage boundary requires an additional point-to-point communication over NeuronLink or EFA, and the interleaved schedule demands careful instruction-level orchestration of send/receive with compute.

Megatron-Core (v0.11+) adds \emph{Sequence Parallelism} (SP, introduced earlier in Megatron-LM v3 circa 2022 \citep{korthikanti2023megatronsp}), \emph{Context Parallelism} (CP, a newer addition for long-context training), and \emph{Expert Parallelism} (EP) for MoE architectures. Each parallelism primitive requires specific communication patterns---all-reduce for DP, all-gather/reduce-scatter for TP, point-to-point for PP---and these patterns must be mapped onto the target hardware's interconnect topology.

\medskip

\noindent\textbf{Layer 2: Fused Kernels.} Performance-critical operations---fused attention (FlashAttention variants), fused MLP layers, fused layer normalization with quantization---are implemented as hand-optimized CUDA kernels, often using NVIDIA's cuBLAS, cuDNN, and Transformer Engine libraries. These kernels exploit NVIDIA-specific hardware features: Tensor Cores for mixed-precision GEMM, shared memory for data reuse, warp-level primitives for fast reductions, and the FP8 data path on Hopper/Blackwell architectures.

\medskip

\noindent\textbf{Layer 3: Communication Overlap.} Megatron's highest-performing configurations overlap computation with communication: TP communication with GEMM execution (\texttt{--tp-comm-overlap}), DP gradient reduction with backward computation (\texttt{--overlap-grad-reduce}), and parameter gathering with forward computation (\texttt{--overlap-param-gather}). These overlaps are orchestrated through CUDA streams and events---fundamentally NVIDIA-specific scheduling primitives.

\medskip

\noindent\textbf{Layer 4: Memory Management.} Distributed checkpointing, activation recomputation, CPU offloading, and gradient accumulation are implemented with assumptions about GPU memory hierarchy (HBM $\to$ L2 $\to$ shared memory $\to$ registers) and CUDA's unified memory model.

\bigskip

The critical observation is that \emph{none of these layers are hardware-portable by default}. Layers~2 and~3 are written in CUDA and depend on NVIDIA-specific hardware features. Layers~1 and~4 use NCCL for communication and CUDA APIs for memory management. Migrating Megatron to Trainium2 therefore requires rewriting or replacing every layer.

% ============================================================
\section{CCCL and the DIA Project: Position in the NVIDIA Fortress}
\label{sec:3.2.cccl}
% ============================================================

Before analyzing the migration path \emph{away} from NVIDIA, we must understand what NVIDIA is building to keep developers \emph{inside} its ecosystem. The CUDA Core Compute Libraries (CCCL) \citep{nvidia2024cccl} and the associated DIA project board (\texttt{github.com/orgs/NVIDIA/projects/6}) are central to this strategy.

\subsection{What CCCL Is}

CCCL unifies three foundational CUDA C++ libraries into a single repository:

\begin{description}
\item[Thrust] -- A parallel algorithms library modeled on the C++ Standard Library, providing \texttt{thrust::sort}, \texttt{thrust::reduce}, \texttt{thrust::transform}, and similar high-level operations that dispatch to optimized GPU implementations.
\item[CUB] -- A library of cooperative GPU primitives at the block and warp level: \texttt{cub::BlockReduce}, \texttt{cub::WarpScan}, \texttt{cub::DeviceRadixSort}. These are the building blocks from which higher-level algorithms are constructed.
\item[libcudacxx] -- A CUDA-compatible implementation of the C++ Standard Library, including \texttt{cuda::std::atomic}, \texttt{cuda::std::tuple}, and memory order primitives that work across host and device code.
\end{description}

The CCCL~3.2 release (as of early 2026; the \texttt{cuda-cccl} Python package reached v0.5.1 in February 2026) introduced modern C++ runtime abstractions (\texttt{cuda::stream}, \texttt{cuda::event}, \texttt{cuda::arch\_traits}), memory resource APIs (\texttt{cuda::buffer}), and new speed-of-light algorithms including \texttt{cub::DeviceTopK}. The unification consolidated APIs from the \texttt{thrust::} and \texttt{cub::} namespaces into \texttt{cuda::} and \texttt{cuda::std::}. Note: the original 2023 unification of Thrust, CUB, and libcudacxx into a single repository should not be confused with a ``CCCL~3.0'' major version---CCCL's major version increments coincide with new CUDA Toolkit major versions, and as of March 2026, CUDA Toolkit~12.x remains current.

CCCL's stated mission is to be ``the Standard C++ Library for CUDA C++''---providing general-purpose, speed-of-light tools so that CUDA developers can focus on their domain problems. Related projects tracked on the DIA project board include NVBench (benchmarking), cuCollections (GPU hash tables), and stdexec (C++26 senders).

\subsection{Where CCCL Sits in the NVIDIA Stack}

The NVIDIA AI training stack can be visualized as a hierarchy. At the bottom lies the hardware (GPU silicon, NVLink/NVSwitch interconnect). Above it, the CUDA driver and runtime. Then CCCL, providing the fundamental parallel primitives. Above CCCL sit the domain-specific libraries: cuBLAS (GEMM), cuDNN (neural network primitives), Transformer Engine (FP8 transformer layers), NCCL (collective communication). Above these sits Megatron-Core, which composes these libraries into distributed training patterns. And at the top sit user-facing frameworks like NeMo and the NeMo Agent Toolkit.

CCCL is thus the \emph{foundation layer}---the equivalent of \texttt{libc} in the Unix world. It does not directly perform matrix multiplication or attention computation, but it provides the parallel reduction, scan, sort, and atomic operations that every higher-level library depends upon. The DIA project board tracks the ongoing development of this foundation.

\subsection{What CCCL Means for the Anthropic Migration}

From the perspective of migrating to Trainium2, CCCL is \emph{precisely what you cannot use}. CCCL is CUDA-only. Its entire purpose is to make CUDA programming more pleasant---it makes the walls of the NVIDIA fortress more comfortable, but they remain fortress walls.

The analogy from Chapter~3.1 is instructive. Recall DeepSeek's approach of writing PTX (NVIDIA's assembly language) to squeeze maximum performance from NVIDIA GPUs. Using CCCL's NVBench to measure performance, using CUB's block-level primitives to structure kernels---these are activities that occur \emph{entirely within} the NVIDIA ecosystem. They are analogous to optimizing the parameters $a$ and $c$ of a linear congruential generator while keeping the modulus $m$ fixed. You may achieve excellent sequences, but you are still bound to that specific modulus.

Anthropic's approach is to change $m$ itself. When you write Trainium2 assembly, you are operating with a different ISA, a different memory hierarchy (no shared memory in the CUDA sense, no warps, no warp-level primitives), a different execution model (VLIW rather than SIMT). The concepts that CCCL abstracts---warp divergence, bank conflicts, occupancy---simply do not exist on Trainium2. Conversely, Trainium2 has its own concerns: VLIW instruction packing, scratchpad memory management via NKI (Neuron Kernel Interface), and the GPSIMD engine's instruction scheduling.

\begin{quote}
\emph{The shared territory between CCCL and Trainium2 kernel programming consists only of the most fundamental optimization principles: instruction-level parallelism, memory latency hiding, and pipeline utilization. The specific mechanisms for achieving these goals are entirely different.}
\end{quote}

This is why the companion document \emph{Anthropic~WantToDoWhat} states that comparing CCCL/NVBench benchmarking methodology with Anthropic's VLIW performance challenge is ``technically strained''---they share a philosophical lineage but inhabit different hardware universes.

% ============================================================
\section{The Migration Map: Megatron Primitives to Trainium2}
\label{sec:3.2.migration}
% ============================================================

We now construct the concrete migration map for each Megatron layer, identifying which NVIDIA-specific components must be replaced and what their Trainium2/Neuron equivalents are (or must be hand-written as).

\subsection{Layer 1: Parallelism $\to$ NxD (Neuron Distributed)}
\label{sec:3.2.migration.parallelism}

Megatron's parallelism primitives (TP, PP, DP, SP, CP, EP) communicate via NCCL, which implements collective operations optimized for NVLink and InfiniBand. On Trainium2, the equivalent is \texttt{neuronx-distributed} (NxD), which implements the same logical operations---all-reduce, all-gather, reduce-scatter---but mapped onto Trainium2's NeuronLink interconnect and EFA (Elastic Fabric Adapter) networking.

The AWS Neuron team has published a reference adaptation of Megatron-LM for Trainium (\texttt{aws-neuron/aws-neuron-reference-for-megatron-lm}), though this has been deprecated in favor of NxD and the NeMo-Megatron reference (\texttt{aws-neuron/neuronx-nemo-megatron}). The key adaptations are:

\begin{description}
\item[Device replacement:] CUDA device references are replaced with PyTorch/XLA devices.
\item[Communication backend:] NCCL is replaced with the PyTorch/XLA distributed backend, which bridges PyTorch's distributed APIs to XLA communication semantics over NeuronLink.
\item[Data loading:] CUDA-specific data loading is replaced with \texttt{MpDeviceLoader}, which overlaps tracing, compilation, and data batch loading.
\item[Optimizers:] CUDA fused optimizers (which use custom CUDA kernels internally) are replaced with generic PyTorch alternatives or Neuron-optimized variants.
\end{description}

\noindent\textbf{Critical limitation:} The Neuron Megatron reference only supports tensor parallelism (degree~8) and data parallelism (up to degree~64). Pipeline parallelism, sequence parallelism, context parallelism, and expert parallelism---all critical for training at Anthropic's scale---require custom implementation. This is precisely the kind of work Anthropic's kernel engineers perform: implementing parallelism strategies at the assembly level where the Neuron compiler's automatic partitioning falls short.

For the Operator RL system (\texttt{operatorRL/src}), this means the agent governance kernel must be aware of which parallelism backend is active. The \texttt{agent\_os} module's policy enforcement and the \texttt{control-plane} scheduler must make hardware-aware decisions about how to partition agent rollout workers across NeuronCores versus GPU SMs.

\subsection{Layer 2: Fused Kernels $\to$ NKI Assembly}
\label{sec:3.2.migration.kernels}

This is the deepest and most technically demanding layer of the migration. Megatron's performance depends on fused CUDA kernels---FlashAttention, fused GEMMs with bias and activation, fused layer normalization with quantization. On NVIDIA hardware, these are implemented using:

\begin{itemize}
\item Tensor Core instructions (\texttt{mma.sync}, \texttt{wgmma} on Hopper)
\item Shared memory for operand staging and accumulation
\item Warp-level primitives (\texttt{\_\_shfl\_sync}, \texttt{\_\_ballot\_sync})
\item Register-level optimizations (register tiling, software pipelining)
\end{itemize}

On Trainium2, \emph{none of these exist}. The equivalent programming model uses the Neuron Kernel Interface (NKI), which as of Neuron SDK 2.27 provides:

\begin{description}
\item[NKI API:] A Python-like DSL (\texttt{nki.language.*}) for writing kernels that execute on Trainium's scalar, vector, and tensor engines. Operations include \texttt{nki.language.add}, \texttt{nki.language.multiply}, \texttt{nki.language.matmul} (mapped to the on-chip matrix engine), and memory operations for explicit DMA between HBM and on-chip SBUF/PSUM.
\item[NKI Library:] Pre-optimized kernels for common operations: Attention (CTE and TKG variants), MLP, Output Projection, QKV, and RMSNorm-Quant. These are the Trainium2 equivalents of cuDNN/Transformer Engine fused kernels.
\item[NKI ISA:] For maximum control, advanced users (such as Anthropic's engineers) can bypass NKI's Python-level API and write instructions directly against the Trainium2 ISA, which is openly documented. This is analogous to writing PTX on NVIDIA hardware, but with a critical difference: the Trainium2 ISA is a \emph{VLIW} (Very Long Instruction Word) architecture, where the programmer (or compiler) explicitly schedules independent operations into each instruction bundle, rather than relying on SIMT hardware to manage thread divergence and reconvergence.
\item[GPSIMD Engine:] Trainium2's \emph{General Purpose SIMD} (GPSIMD) engine is the scalar/vector processing unit responsible for non-tensor operations: index computation, conditional branching, loop control, address generation, and arbitrary data-dependent control flow. While the matrix engine handles the heavy-compute GEMM/attention workloads, the GPSIMD engine handles everything the matrix engine cannot: computing softmax denominators, generating reduction indices, evaluating activation functions element-wise, and orchestrating the DMA engine's scatter-gather patterns. Anthropic's kernel engineers use GPSIMD for the ``glue'' logic that connects tensor operations---the scheduling overhead that, on NVIDIA hardware, would be handled by warp-level control flow and predicated execution. Efficient GPSIMD utilization is critical: in a fused attention kernel, up to 15--20\% of total cycles may be spent on GPSIMD operations (softmax scaling, mask application, causal masking logic), and poor GPSIMD scheduling can bottleneck an otherwise well-optimized tensor pipeline.
\end{description}

The migration of each critical kernel type proceeds as follows:

\medskip

\noindent\textbf{GEMM (Matrix Multiply):} On NVIDIA, this is cuBLAS dispatching to Tensor Core \texttt{mma} instructions. On Trainium2, the on-chip matrix multiply engine (MatMul Engine) performs the equivalent operation, but tiling, data layout, and accumulation are managed through explicit \textbf{SBUF} (State Buffer: the on-chip scratchpad memory for staging input operands, analogous to CUDA shared memory) and \textbf{PSUM} (Partial Sum: the accumulator scratchpad for output operands, where intermediate matrix multiply results are accumulated before being written back to HBM) orchestration rather than shared memory and Tensor Core abstractions.

\medskip

\noindent\textbf{Attention:} FlashAttention \citep{dao2022flashattention,dao2023flashattention2} on NVIDIA uses shared memory to tile Q, K, V matrices and compute softmax in a single fused kernel, avoiding materialization of the full attention matrix in HBM. On Trainium2, the NKI Library provides Attention CTE (Compile-Time Execution) and Attention TKG variants that achieve the same algorithmic optimization but map it onto Trainium's scratchpad architecture. Anthropic's engineers have written custom attention assembly that further optimizes these patterns for Claude's specific model dimensions.

\medskip

\noindent\textbf{Communication Primitives (AllReduce, AllGather):} On NVIDIA, NCCL uses GPU-initiated communication via NVLink and InfiniBand RDMA. On Trainium2, collective communication uses the NeuronLink interconnect with DMA engines that can operate independently of the compute engines. This architectural difference enables a different style of compute-communication overlap: rather than using CUDA streams to interleave kernel launches with NCCL calls, Trainium2 can use its DMA engines to perform data transfers \emph{concurrently with} tensor engine computation at the hardware level.

\subsection{Layer 3: Communication Overlap $\to$ DMA Engine Orchestration}
\label{sec:3.2.migration.overlap}

Megatron's communication overlap patterns (\texttt{--tp-comm-overlap}, \texttt{--overlap-grad-reduce}) rely on CUDA streams and events to schedule independent work concurrently. On Trainium2, the equivalent capability comes from the chip's DMA engines, which can perform HBM-to-HBM and NeuronLink transfers independently of the compute pipeline.

For Agent Lightning's training loop (\texttt{agentlightning/trainer}), this means the \texttt{verl}-based training orchestration must be adapted to exploit Trainium2's DMA-based overlap rather than stream-based overlap. The reward computation (\texttt{agentlightning/reward.py}) and trajectory storage (\texttt{agentlightning/store}) can remain hardware-agnostic, but the gradient synchronization in the training inner loop requires platform-specific code paths.

\subsection{Layer 4: Memory Management $\to$ Neuron Runtime}
\label{sec:3.2.migration.memory}

Megatron's activation recomputation, gradient accumulation, and distributed checkpointing assume CUDA's memory model. On Trainium2, the Neuron Runtime manages HBM allocation and NeuronCore-local memory. The key differences are:

\begin{description}
\item[No unified memory:] Trainium2 does not support CUDA-style unified memory or managed memory. All data movement between host and device must be explicit.
\item[Serialized checkpointing:] Neuron's checkpointing uses serialized tensor save/load with staggered writes to alleviate memory pressure---a different approach from Megatron's distributed checkpoint writer.
\item[Compilation overhead:] The XLA compilation model means that the first execution of a new graph incurs compilation cost. Neuron's ahead-of-time parallel compilation mitigates this, but it represents a fundamentally different execution model from CUDA's eager kernel launch.
\end{description}

% ============================================================
\section{The Agentic RL Integration: Self-Deploying Systems on Heterogeneous Hardware}
\label{sec:3.2.agenticrl}
% ============================================================

We now connect the migration analysis to the central theme of this volume: agentic RL systems that self-deploy, gather environment feedback, and self-evolve. The key insight is that an agentic system like Claude Code running autonomously must be capable of \emph{selecting its own hardware backend} based on availability, cost, and performance---and the kernel stack must support this dynamic selection.

\begin{figure}[ht]
\centering
\small
\begin{verbatim}
+================================================================+
|                    OPERATOR RL (Governance)                     |
|  +------------------+  +------------------+  +---------------+ |
|  | SCAK (Layer 4)   |  | Policy Engine(3) |  | IATP (Layer 2)| |
|  | Self-Correction  |  | Syscall Check    |  | Trust Boundary| |
|  +--------+---------+  +--------+---------+  +-------+-------+ |
|           |                     |                     |         |
|           v                     v                     v         |
|  +-------------------+  +------------------+  +--------------+ |
|  | Primitives (L1)   |  | Control Plane    |  | Nexus        | |
|  | action/obs/reward |  | Scheduler        |  | Multi-Agent  | |
|  +-------------------+  +------------------+  +--------------+ |
+===========================|=====================================+
                            |  Kernel Bridge (PR #478)
                            v
+================================================================+
|                 AGENT LIGHTNING (Training)                      |
|  +----------+  +-----------+  +--------+  +-----------------+ |
|  | Runner   |  | Tracer    |  | Store  |  | Daemon          | |
|  | (rollout)|  | (capture) |  | (7 bk) |  | v0=HTTP/v1=Store| |
|  +----+-----+  +-----+-----+  +---+----+  +--------+--------+ |
|       |              |             |                 |          |
|       v              v             v                 v          |
|  +----------+  +----------------+  +---------------------------+|
|  | Adapter  |  | trace_agg     |  | VERL Backend               ||
|  | LangChain|  | transition/   |  | GRPO / PPO                 ||
|  | CrewAI   |  | trajectory    |  | Distributed Training       ||
|  +----------+  +----------------+  +---------------------------+|
+===========================|=====================================+
                            |  Hardware Abstraction Boundary
                            v
+================================================================+
|              HARDWARE KERNEL LAYER (per-platform)              |
|  +-------------------+ +-------------------+ +---------------+ |
|  | NVIDIA GPU        | | Google TPU        | | AWS Trainium2 | |
|  | CUDA/cuBLAS/NCCL  | | XLA/JAX           | | NKI/NxD/ISA   | |
|  | Megatron-Core     | | Pathways          | | Hand-written  | |
|  |  (mature)         | |  (internal)       | | assembly      | |
|  +-------------------+ +-------------------+ +---------------+ |
+================================================================+
\end{verbatim}
\caption{System architecture: Operator RL (governance) $\to$ Agent Lightning (training) $\to$ Hardware Kernel Layer (per-platform). The kernel bridge (PR~\#478, commit \texttt{49bf9cd9}) connects governance syscalls to the training loop. The hardware abstraction boundary separates platform-agnostic orchestration from platform-specific kernel implementations. The three hardware columns represent the asymmetric deployment risk analyzed in \S\ref{sec:3.1}: mature (GPU), internal (TPU), and youngest (Trainium2).}
\label{fig:system-arch-ascii}
\end{figure}

\subsection{The Operator RL Architecture on Multi-Platform Kernels}

The Operator RL system (\texttt{operatorRL}) provides the governance layer for autonomous agents. Its kernel architecture (\texttt{src/agent\_os}) acts as an operating system for agents, enforcing policies, managing resources, and scheduling tasks.

\medskip
\noindent\textbf{The Four Kernel Invariants.} The Agent~OS stateless kernel (see \texttt{operatorRL/docs/kernel-internals.md}) enforces four invariants that distinguish it from library-based agent frameworks such as LangChain or CrewAI:

\begin{description}
\item[\textsc{Invariant~1: No Bypass.}] All agent I/O passes through kernel-space syscalls (\texttt{SYS\_EXEC}, \texttt{SYS\_WRITE}, etc.). No agent action reaches the environment without first transiting the policy engine. In the terminology of \S\ref{sec:3.2.migration.kernels}, this is the governance equivalent of requiring every tensor operation to pass through the NKI runtime rather than directly accessing HBM.

\item[\textsc{Invariant~2: Fail Closed.}] Unknown or unrecognized actions are denied by default. The kernel emits \texttt{SIGKILL} (non-maskable) on any policy violation. This is analogous to a VLIW instruction bundle where an unscheduled operation slot is filled with a NOP rather than left undefined---fail-safe by construction.

\item[\textsc{Invariant~3: Audit Everything.}] The Flight Recorder logs every execution decision, every signal dispatch, and every policy check result. For multi-platform agentic RL, this audit trail is the only reliable mechanism for post-hoc debugging across the Cartesian product of hardware backends (\S\ref{sec:3.2.critical}).

\item[\textsc{Invariant~4: Crash Isolation.}] A user-space agent crash (e.g., a rollout worker segfault on Trainium2 due to a DMA misconfiguration) does not crash the kernel. The governance layer remains operational and can restart the agent on an alternative backend.
\end{description}

\noindent These invariants hold regardless of which hardware backend executes the agent's computational workload. The kernel itself is hardware-agnostic; the hardware-specific code lives entirely in the training loop (Agent Lightning) and the kernel migration layers (\S\ref{sec:3.2.migration}).

When integrated with Agent Lightning's training infrastructure (\texttt{agentlightning}), the combined system forms a self-evolving training loop:

\begin{enumerate}
\item \textbf{Agent rollout} (Agent Lightning \texttt{runner}): The agent executes in a live environment, generating trajectories. The rollout engine must run on \emph{whatever hardware is available}---GPU, TPU, or Trainium2---and the kernel stack must provide equivalent throughput on each.

\item \textbf{Reward computation} (Agent Lightning \texttt{reward.py} + Operator RL \texttt{integrations/agent\_lightning/reward.py}): Trajectory rewards combine a task-completion signal $R_{\mathrm{base}}$ with governance penalties from the \textsc{PolicyReward} class. The integration bridge's reward formula is:
\[
R_{\mathrm{final}} = \begin{cases}
R_{\mathrm{base}} + \displaystyle\sum_{v \in \mathcal{V}} \phi(v) + b_{\mathrm{clean}} \cdot \mathbf{1}[\mathcal{V} = \emptyset] & \text{(additive mode)} \\[6pt]
R_{\mathrm{base}} \cdot \alpha & \text{(multiplicative mode, any } v \in \mathcal{V}\text{)}
\end{cases}
\]
where $\phi(v)$ maps violation severity to penalties (critical: $-100$, high: $-50$, medium: $-10$, low: $-1$), $b_{\mathrm{clean}} = 5$ is a compliance bonus, and $\alpha = 0.5$ is the multiplicative discount factor. The reward is clamped to $[R_{\min}, R_{\max}]$ (default $[-100, 100]$). This is \emph{not} a standard inference workload---it interleaves kernel policy evaluation with scalar reward computation, and on Trainium2, the policy evaluation step (which invokes the governance kernel's rule engine) runs on the GPSIMD engine while the GEMM-heavy forward pass occupies the matrix engine, enabling overlap.

\item \textbf{Policy update} (Agent Lightning \texttt{trainer}, with \texttt{verl} backend): The agent's policy parameters are updated using RL algorithms (GRPO \citep{shao2024deepseekmath} or PPO \citep{schulman2017ppo} via VERL). This is a distributed training step that requires all the Megatron-equivalent parallelism primitives discussed in Section~\ref{sec:3.2.migration.parallelism}.

\item \textbf{Self-deployment} (Operator RL \texttt{control-plane}): The updated model is deployed back into the environment. The governance layer ensures safety constraints are maintained during autonomous operation.

\item \textbf{Iteration}: The loop repeats, with the agent continuously improving through interaction with its environment.
\end{enumerate}

For this loop to function on Trainium2 specifically, each step requires the kernel migrations described in Section~\ref{sec:3.2.migration}. The critical path is Step~3 (policy update), where the full distributed training stack---parallelism, fused kernels, communication overlap, memory management---must operate at near-peak hardware utilization.

\medskip
\noindent\textbf{The Integration Bridge: Four Adapter Components.} The directory \texttt{operatorRL/src/agent\_os/integrations/agent\_lightning/} contains the concrete bridge between the governance kernel and the RL training loop. Four Python modules implement the bridge; each maps a governance concept to an Agent Lightning abstraction.

\begin{description}
\item[\texttt{runner.py}: \textsc{GovernedRunner}.] Wraps Agent Lightning's \texttt{Runner} interface with kernel policy enforcement. The key mechanism is the \texttt{step()} coroutine: before every agent action reaches the environment, the kernel's \texttt{execute\_async()} interposes a policy check. If a violation is detected, the runner records a \texttt{PolicyViolation} object annotated with severity (\texttt{critical}, \texttt{high}, \texttt{medium}, \texttt{low}) and type (\texttt{BLOCKED}, \texttt{MODIFIED}, \texttt{WARNED}, \texttt{SIGNAL\_SENT}). The rollout result is wrapped in a \texttt{GovernedRollout} dataclass that carries both the task output and the violation metadata. Critically, \texttt{GovernedRunner} follows Agent Lightning's runner lifecycle: \texttt{init()} $\to$ \texttt{init\_worker(worker\_id, store)} $\to$ \texttt{step()} / \texttt{iter()} $\to$ \texttt{teardown\_worker()} $\to$ \texttt{teardown()}. This lifecycle alignment is what makes the governance layer transparent to the VERL backend---the trainer sees a standard runner that happens to produce richer metadata.

\item[\texttt{reward.py}: \textsc{PolicyReward}.] Converts governance violations into RL training signals. The reward computation is $R_{\mathrm{final}} = R_{\mathrm{base}} + \sum_{v \in \mathcal{V}} \phi(v) + \mathbf{1}[\mathcal{V} = \emptyset] \cdot b_{\mathrm{clean}}$, where $R_{\mathrm{base}}$ is the task reward, $\phi(v)$ maps violation severity to a penalty (default: critical $= -100$, high $= -50$, medium $= -10$, low $= -1$), and $b_{\mathrm{clean}} = 5$ is a bonus for zero-violation execution. An alternative multiplicative mode sets $R_{\mathrm{final}} = R_{\mathrm{base}} \cdot \alpha$ (default $\alpha = 0.5$) on any violation. The \texttt{CompositeReward} class further supports multi-dimensional rewards with configurable weights---addressing the multi-objective tension identified in Issue~\#490.

\item[\texttt{emitter.py}: \textsc{FlightRecorderEmitter}.] Adapts the kernel's Flight Recorder audit log to Agent Lightning's \texttt{LightningStore} span format. Each Flight Recorder entry (policy check, signal dispatch, tool call) is converted to a \texttt{LightningSpan} with structured attributes (e.g., \texttt{agent\_os.policy\_violated}, \texttt{agent\_os.signal\_type}). The emitter supports both batch export (\texttt{emit\_to\_store()}) and async streaming. For the validation tests of \S\ref{sec:3.3}, the \texttt{get\_violation\_summary()} method provides the per-policy violation rates needed by the $\chi^2$ test on governance compliance distribution.

\item[\texttt{environment.py}: \textsc{GovernedEnvironment}.] A Gym-compatible environment wrapper that converts kernel governance into standard RL signals: $(\mathit{state}, \mathit{reward}, \mathit{terminated}, \mathit{truncated}, \mathit{info})$. The environment terminates the episode on critical violations (configurable via \texttt{terminate\_on\_critical}), applies per-step efficiency penalties (default $-0.1$), and tracks compliance metrics across episodes. On Trainium2, the environment's \texttt{step()} function is the outermost loop of the training pipeline---its per-call overhead (kernel policy check + Flight Recorder write + violation penalty computation) must be negligible compared to the agent's forward pass latency, which is dominated by the NKI attention kernel execution time analyzed in \S\ref{sec:3.2.migration.kernels}.
\end{description}

\noindent The combined effect: Agent Lightning trains agents that are simultaneously optimized for task performance \emph{and} governance compliance, because the RL reward signal encodes both. Table~\ref{tab:governance-training} (from the integration README) quantifies this: agents trained with \textsc{GovernedRunner} achieve 0.0\% policy violations while maintaining higher task accuracy (79.2\% vs 76.4\%) than ungoverned baselines---the governance signal acts as an implicit regularizer that prevents reward hacking via policy-violating shortcuts.

\medskip
\noindent\textbf{Signal Dispatch Table.} The Agent~OS kernel communicates with agents through a POSIX-inspired signal mechanism (see \texttt{operatorRL/packages/control-plane/src/\allowbreak agent\_control\_plane/signals.py}). Table~\ref{tab:signals} defines the signal set.

\begin{table}[h]
\centering
\caption{Agent~OS kernel signals. Non-maskable signals enforce governance invariants and cannot be intercepted by user-space agent code.}
\label{tab:signals}
\begin{tabular}{@{}lclcc@{}}
\toprule
\textbf{Signal} & \textbf{Value} & \textbf{Description} & \textbf{Maskable} & \textbf{Recovery} \\
\midrule
\texttt{SIGSTOP}   & 1 & Pause agent execution         & Yes & \texttt{SIGCONT} \\
\texttt{SIGCONT}   & 2 & Resume agent execution         & Yes & --- \\
\texttt{SIGINT}    & 3 & Graceful interrupt             & Yes & Restart \\
\texttt{SIGKILL}   & 4 & Immediate termination          & \textbf{No}  & New agent \\
\texttt{SIGTERM}   & 5 & Request shutdown               & Yes & Cleanup \\
\texttt{SIGPOLICY} & 8 & Policy violation (Invariant~2) & \textbf{No}  & None \\
\texttt{SIGTRUST}  & 9 & Trust violation (IATP Layer~2) & \textbf{No}  & None \\
\bottomrule
\end{tabular}
\end{table}

\noindent The non-maskable signals (\texttt{SIGKILL}, \texttt{SIGPOLICY}, \texttt{SIGTRUST}) are the enforcement arm of the kernel invariants. When the policy engine detects a violation---whether a blocked action, a blocked pattern match in parameters, or an inter-agent trust boundary breach---the kernel emits the appropriate signal, logs the event to the Flight Recorder, and terminates the agent. In a multi-platform training loop, this mechanism applies identically regardless of whether the agent's rollout was executing on GPU, TPU, or Trainium2: the signal dispatch operates at the governance layer, above the hardware abstraction boundary.

\subsection{The Agent Lightning Daemon: Two Execution Modes}
\label{sec:3.2.daemon}

The \texttt{AgentModeDaemon} class (see \texttt{agentlightning/verl/daemon.py}, PR~\#407, commit \texttt{1d199b21}) is the bridge between Agent Lightning's agent rollout infrastructure and the VERL-based distributed training backend. It operates in two modes, selectable at initialization via the \texttt{mode} parameter:

\begin{description}
\item[\texttt{mode="v0"} (Server Mode):] The daemon runs a Flask-based HTTP server (\texttt{AgentLightningServer}) on a fixed port. Agent rollout workers communicate with the training loop via REST endpoints. An LLM proxy server runs on a separate port, intercepting model inference requests and routing them to backend vLLM/SGLang instances. This mode is suitable for single-node or small-cluster deployments where network topology is simple.

\item[\texttt{mode="v1"} (Store Mode):] The daemon bypasses the HTTP server entirely and communicates through a \texttt{LightningStore} backend (with 7~implementations: memory, SQLite, MongoDB, client-server, collection-based, threaded, and base). An internal asyncio event loop runs on a daemon thread for asynchronous store operations. The \texttt{LLMProxy} and \texttt{TraceToTripletBase} adapter handle trace-to-training-data conversion. This mode is designed for large-scale distributed deployments where the store provides coordination across many training and rollout workers.
\end{description}

\noindent The daemon's \texttt{trace\_aggregator} parameter (added in PR~\#134, commit \texttt{4235731a}) controls whether rollout data is aggregated at the \emph{transition} level (individual state-action-reward tuples) or the \emph{trajectory} level (complete multi-turn episodes). This distinction is critical for agentic RL: transition-level aggregation enables per-step credit assignment but loses episode-level context; trajectory-level aggregation preserves the full decision trace but requires more memory and complicates gradient computation. The choice directly affects how the training loop interacts with the hardware memory hierarchy described in \S\ref{sec:3.2.migration.memory}---on Trainium2, trajectory-level aggregation may exceed the XLA runtime's memory planner capacity for long episodes, triggering the OOM failures documented in Issue~\#107.

For the hardware migration, the daemon itself is hardware-agnostic (pure Python orchestration). The hardware-specific code paths lie in the \texttt{verl} training backend that the daemon feeds data into, and in the LLM proxy's routing to platform-specific inference servers. This separation of concerns---governance (Operator~RL kernel) $\to$ orchestration (daemon) $\to$ computation (VERL on GPU/TPU/Trainium2)---is the architectural spine of the entire system.

\subsection{The OperatorRL--Agent Lightning Integration Bridge}
\label{sec:3.2.bridge}

The formal connection between the governance layer (Operator~RL) and the training loop (Agent Lightning) was established in PR~\#478 (commit \texttt{49bf9cd9}, ``Agent-OS Integration: Kernel-Level Safety for RL Training'').  The bridge operates as follows:

\begin{enumerate}
\item \textbf{Syscall Interception.} Before each rollout step, the Agent Lightning \texttt{runner} issues a kernel syscall through the Agent~OS stateless kernel (\texttt{operatorRL/src/agent\_os/stateless.py}). The syscall contains the proposed action, its parameters, and the current \texttt{ExecutionContext} (agent ID, active policies, history).

\item \textbf{Policy Enforcement.} The kernel's policy engine evaluates the request against all active policies---checking blocked actions, blocked patterns (via regex over serialized parameters), and any custom constraints (e.g., the ``banking'' policy: block transfers exceeding a threshold). If any check fails, the kernel emits \texttt{SIGKILL} and the rollout step is aborted.

\item \textbf{Context Propagation.} On success, the kernel returns an updated \texttt{ExecutionContext} with the action appended to the history. This context is passed back to the daemon, which incorporates it into the trace data fed to VERL for gradient computation. The trace\_aggregator determines whether this context propagates at transition or trajectory granularity.

\item \textbf{VFS State.} The agent's episodic memory, working memory, and policy state are exposed through the Agent~OS Virtual File System (VFS) at mount points \texttt{/agent/\{id\}/mem/\{working,episodic,semantic\}} and \texttt{/agent/\{id\}/policy/}. The VFS is backed by pluggable storage (in-memory for testing, Redis for distributed production, DynamoDB for serverless)---mirroring the LightningStore's multiple backend strategy.
\end{enumerate}

\noindent The bridge ensures that the governance invariants (\S\ref{sec:3.2.agenticrl}) hold throughout the self-evolving training loop, even as the underlying computation migrates across hardware backends. On Trainium2, the governance checks add negligible latency (they execute in Python on the CPU host), while the kernel migration costs are entirely in the VERL training step where NKI assembly and NxD communication primitives replace their CUDA equivalents.

\subsection{Why Megatron's Techniques Must Be Migrated, Not Abandoned}

One might ask: why not simply use Trainium2's native Neuron SDK tools (\texttt{neuronx-distributed}, \texttt{neuronx-nemo-megatron}) and avoid the migration complexity? The answer lies in the performance gap between compiler-generated and hand-optimized kernels, which is precisely the theme of Chapter~3.1's discussion of ``apparently optimal'' versus truly optimal kernel configurations.

The Neuron compiler's XLA-based optimization pipeline produces kernels that are ``apparently optimal''---they appear to utilize the hardware efficiently based on the compiler's cost model. But just as Algorithm~K in Chapter~3.1 appeared to produce an inexhaustible supply of convincingly optimal configurations yet immediately converged to a fixed point, compiler-generated kernels often leave significant performance on the table.

Anthropic's engineers have demonstrated that hand-written Trainium2 assembly can substantially outperform compiler-generated code for the specific kernels that dominate training and inference time (GEMM, attention, collective communication). This is the same insight that led DeepSeek to write PTX for NVIDIA GPUs---but applied to a different hardware target.

The moral, echoing Chapter~3.1: \emph{optimal hardware kernel configurations are not produced by randomly chosen methods; certain theory should be used.}

% ============================================================
\section{Critical Analysis: Engineering Cost and Degenerate Cases}
\label{sec:3.2.critical}
% ============================================================

\subsection{The Cartesian Product of Debug Spaces}

Maintaining optimized kernel stacks across three platforms (GPU/TPU/Trainium) simultaneously is extraordinarily expensive. If each platform has $k$ critical kernels and each kernel has $d$ potential failure modes, the total debug space is $O(k + d)$ in the best case (failures are platform-independent and additive, so debugging one platform transfers to others) and $O(3 \cdot k \cdot d)$ in the worst case (each platform--kernel--failure-mode triple is unique, forming a Cartesian product). In practice, the worst case is closer to reality: a bank conflict on GPU, a tiling inefficiency on TPU, and a VLIW scheduling stall on Trainium are completely different bugs requiring different debugging tools and expertise.

This is the ``Cartesian product'' cost described in \emph{Anthropic~WantToDoWhat}. The justification for bearing this cost is strategic: vendor independence is worth the engineering investment because it provides negotiating leverage and protects against supply chain disruptions.

\subsection{Degenerate Configurations in Cross-Platform Auto-Tuning}

Consider an auto-tuner that searches for optimal kernel configurations across platforms. On each platform, the search space has a different structure: GPU configurations involve tile sizes, thread block dimensions, and shared memory allocation; TPU configurations involve XLA layout assignments and fusion decisions; Trainium configurations involve VLIW instruction packing and scratchpad allocation.

A configuration that achieves 95\% utilization on GPU and TPU provides \emph{no information} about what will achieve 95\% on Trainium---the hardware architectures are sufficiently different that the configuration spaces are essentially independent. This is analogous to Exercise~6 of Chapter~3.1: the function $f(\mathbf{x})$ that maps one configuration to the next is different on each platform, so the cycle structure (period, pre-period length) is also different.

\subsection{Where CCCL's DIA Board Fits---And Does Not Fit}

The CCCL project board (DIA) tracks NVIDIA's efforts to build a ``Standard Library for CUDA C++.'' Within the NVIDIA ecosystem, this is enormously valuable: CCCL provides the vetted, high-performance building blocks (parallel sort, reduce, scan, atomics) that every CUDA developer needs. Projects like NVBench (benchmarking) and cuCollections (GPU data structures) extend this foundation.

For our purposes---building hardware-portable agentic RL systems---CCCL/DIA occupies a specific and limited position:

\begin{enumerate}
\item \textbf{As reference implementation:} CCCL's algorithms (CUB's \texttt{BlockReduce}, \texttt{DeviceRadixSort}, \texttt{DeviceTopK}) represent the state of the art for GPU-optimized parallel primitives. When writing the Trainium2 equivalent of these algorithms, CCCL serves as the performance benchmark to beat---the ``gold standard'' configuration sequence against which new hardware's kernel quality is measured.

\item \textbf{As non-portable dependency:} Any code that uses CCCL directly is locked to NVIDIA GPUs. For a multi-platform agentic system, CCCL code must be isolated behind a hardware abstraction layer and replaced with platform-specific implementations on TPU and Trainium2.

\item \textbf{As ecosystem anchor:} CCCL is part of NVIDIA's strategy to make the CUDA ecosystem so comprehensive and convenient that developers never want to leave. The DIA board's tracking of features like \texttt{cuda::stream}, \texttt{cuda::event}, and C++26 senders integration shows NVIDIA investing in making CUDA C++ feel like modern C++---a powerful retention mechanism.
\end{enumerate}

In the language of \emph{Anthropic~WantToDoWhat}: CCCL makes the NVIDIA prison cell more luxurious. It does not help you escape.

% ============================================================
\section{The Open-Source Counterstrategy: XLA, JAX, and NKI}
\label{sec:3.2.opensource}
% ============================================================

AWS's counterstrategy to CCCL is a three-pronged open-source effort:

\begin{description}
\item[XLA Graph Compiler:] Open-source compiler that lowers high-level computation graphs to hardware-specific instructions. Supports GPU (via CUDA), TPU, and Trainium (via Neuron). This is the ``portable'' layer---but as discussed, portability comes at a performance cost relative to hand-written kernels.

\item[JAX Software Stack:] Functional transformation framework (grad, vmap, pmap, jit) built on XLA. Provides a hardware-agnostic programming model that compiles to any XLA backend. AWS is investing in JAX support on Trainium to attract researchers from the TPU/JAX ecosystem.

\item[NKI (Neuron Kernel Interface):] The ``escape hatch'' for when XLA's auto-generated kernels are insufficient. NKI allows writing Trainium-native kernels in a Python-like DSL, with the option to drop down to ISA-level assembly for maximum control. The NKI Library provides pre-optimized kernels for the most critical operations.
\end{description}

The parallel to NVIDIA's stack is:
\begin{center}
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{NVIDIA Stack} & \textbf{AWS/Trainium Stack} \\
\midrule
CUDA Runtime & Neuron Runtime \\
CCCL (Thrust/CUB/libcudacxx) & NKI Library (pre-built kernels) \\
cuBLAS/cuDNN/Transformer Engine & NKI custom kernels / XLA auto-tuned \\
NCCL & NeuronLink collectives (via NxD) \\
Megatron-Core & neuronx-distributed (NxD) / NxD Inference \\
NeMo Framework & neuronx-nemo-megatron (deprecated $\to$ NxDI) \\
PTX (manual assembly) & Trainium2 ISA (manual assembly) \\
Nsight Systems / Nsight Compute & Neuron Profile (\texttt{neuron-profile}) \\
cuda-gdb / compute-sanitizer & \texttt{neuron-top}, \texttt{neuron-monitor} \\
\bottomrule
\end{tabular}
\end{center}

The key asymmetry: NVIDIA's stack is mature, deeply integrated, and supported by millions of developers. AWS's stack is younger, evolving rapidly (the Neuron SDK has released 27+ major versions in quick succession, with breaking changes between versions), and supported by a smaller but growing community. Anthropic's bet is that for the narrow vertical of AI training/inference, a small team of assembly-level experts can achieve equivalent or superior performance on Trainium2, provided they have direct ISA access---which Trainium2's open architecture grants.

% ============================================================
\section{Validation Tests for Agentic System Quality}
\label{sec:3.3}
% ============================================================

Our primary goal is to produce agentic RL systems whose training runs, kernel configurations, and agent behavior exhibit the properties we desire. In \S\ref{sec:3.2} we established how to construct hardware-portable kernel stacks with long effective lifetimes before degeneration; this is an important criterion, but it does not guarantee that the resulting agent behavior is satisfactory for deployment. How, then, does one determine whether an agentic training run is producing sufficiently high-quality results?

If a human engineer examines a sequence of training loss values, reward curves, or kernel utilization metrics, the chance of correctly identifying subtle pathologies is small. Engineers tend to overlook phenomena that \emph{appear} normal---such as a reward plateau that is actually a convergent cycle (analogous to the cycle degeneration of \S\ref{sec:3.1})---while flagging apparent anomalies that are statistically expected. Therefore, we must apply systematic, unbiased quantitative tests.

Statistical theory provides us with quantitative measures of training quality. There are innumerably many conceivable tests; we shall discuss those that have proven most useful, most informative, and most amenable to automated computation within the training loop itself.

If a training run passes tests $T_1, T_2, \ldots, T_n$, we cannot guarantee it will not fail under test $T_{n+1}$; but each additional passing test increases our confidence. In practice, we apply five or six different types of validation tests to each training run, and if all are satisfactorily passed, we consider the run acceptable---innocent until proven guilty.

Two types of tests must be distinguished: \emph{empirical tests}, where the training infrastructure (Agent Lightning's \texttt{trace\_aggregator} and \texttt{store}) collects runtime statistics and computes validation metrics; and \emph{theoretical tests}, where the mathematical structure of the kernel configuration recurrence (\S\ref{sec:3.2}) and the RL algorithm convergence guarantees are used to predict expected behavior.

\subsection{General Test Methods for Agent Training Quality}
\label{sec:3.3.1}

\noindent\textbf{A. The $\chi^2$ Test for Utilization Uniformity.}\quad The $\chi^2$ test is perhaps the most fundamental validation tool. In the agentic RL context, we apply it to three distributions:

\begin{description}
\item[(i) NeuronCore Utilization Distribution.] If the training loop distributes work across $k$~NeuronCores, each core should process approximately $n/k$ of the total $n$~training steps. Let $Y_s$ be the number of steps processed by core~$s$, and $p_s = 1/k$ be the expected fraction under uniform distribution. The $\chi^2$ statistic is
\[
V = \sum_{s=1}^{k} \frac{(Y_s - n/k)^2}{n/k}
\]
with $\nu = k-1$ degrees of freedom. A value of $V$ that falls below the $p=1\%$ percentile of the $\chi^2$ distribution indicates suspiciously uniform utilization (possible measurement error); a value above the $p=99\%$ percentile indicates a load imbalance that may arise from heterogeneous kernel performance across NeuronCores.

\item[(ii) Reward Value Distribution.] Across $n$~rollout episodes, partition the reward range into $k$~equal-width bins. The distribution of observed rewards into these bins should match the expected distribution under the current policy. A significant $\chi^2$ deviation indicates either reward model miscalibration or exploration collapse (Issue~\#490).

\item[(iii) Gradient Norm Distribution.] Across training steps, the distribution of per-layer gradient norms should follow a pattern consistent with the optimizer's expected behavior. Bimodal gradient distributions, detectable via $\chi^2$ on quantile bins, indicate that some layers are training while others are stalled---a symptom of mixed-precision misconfiguration that manifests differently on CUDA (NaN propagation through FP16 underflow) and Trainium2 (BF16 exponent range behavior in the matrix engine).
\end{description}

\noindent The $\chi^2$ table (reproduced from TAOCP Vol.~2, Table~3.3.1-1) provides critical values for interpreting $V$:

\begin{center}
\begin{tabular}{@{}crrrrrrr@{}}
\toprule
$\nu$ & $p=1\%$ & $p=5\%$ & $p=25\%$ & $p=50\%$ & $p=75\%$ & $p=95\%$ & $p=99\%$ \\
\midrule
1  & 0.00016 & 0.00393 & 0.1015 & 0.4549 & 1.323 & 3.841  & 6.635 \\
2  & 0.02010 & 0.1026  & 0.5754 & 1.386  & 2.773 & 5.991  & 9.210 \\
5  & 0.5543  & 1.1455  & 2.675  & 4.351  & 6.626 & 11.07  & 15.09 \\
10 & 2.558   & 3.940   & 6.737  & 9.342  & 12.55 & 18.31  & 23.21 \\
20 & 8.260   & 10.85   & 15.45  & 19.34  & 23.83 & 31.41  & 37.57 \\
\bottomrule
\end{tabular}
\end{center}

\noindent For a Trainium2 cluster with $k=64$ NeuronCores ($\nu=63$), the 95th percentile of $\chi^2$ is approximately $82.5$. If $V > 82.5$, the load imbalance is statistically significant at the 5\% level and warrants investigation of the NxD parallelism configuration.

\subsection{Empirical Tests for Operational Health}
\label{sec:3.3.2}

The classical randomness tests from TAOCP~\S3.3.2 (run test, gap test, serial test, poker test) each have a precise analog in agentic system validation:

\begin{description}
\item[Run Test $\to$ Rollout Length Stability.] In a sequence of rollout episode lengths $\ell_1, \ell_2, \ldots, \ell_n$, a ``run up'' is a maximal subsequence $\ell_i < \ell_{i+1} < \cdots < \ell_{i+r}$. Under stable training, the distribution of run lengths should match the theoretical distribution for independent observations. Excessively long runs indicate reward drift; excessively short runs indicate policy oscillation. In Agent Lightning, the \texttt{store} backends record episode lengths, and the \texttt{trace\_aggregator} at trajectory level preserves the sequential ordering needed for this test.

\item[Gap Test $\to$ Inter-Failure Interval Analysis.] If kernel failures (OOM events, NKI compilation errors, DMA timeouts) occur at steps $t_1, t_2, \ldots$, the gaps $g_i = t_{i+1} - t_i$ should follow a geometric distribution under the assumption of independent, memoryless failures. Clustering of failures (gaps shorter than expected) indicates a systematic hardware issue---on Trainium2, this may signal SBUF fragmentation or NeuronLink congestion. Issue~\#107 (server OOM during training) and Issue~\#438 (memory leak) would both be detectable via gap test anomalies.

\item[Serial Test $\to$ Consecutive Action Correlation.] In the agent's action sequence $a_1, a_2, \ldots, a_n$, the serial test examines pairs $(a_i, a_{i+1})$. Under a well-trained policy with sufficient exploration, the joint distribution of consecutive actions should approximate the product of marginals (independence). High serial correlation indicates the agent is stuck in a repetitive pattern---precisely the ``exploration collapse'' of Issue~\#490. The \texttt{trace\_aggregator} at transition level provides the pair data needed for this test.

\item[Poker Test $\to$ Action Diversity.] Partition the action sequence into groups of $k$~consecutive actions and classify each group by its pattern (all same, all different, one pair, etc.). The poker test compares the observed pattern frequencies against their theoretical probabilities. Low action diversity---detectable as an excess of ``all same'' or ``one pair'' patterns---indicates insufficient exploration and may signal that the GRPO or PPO advantage estimates have collapsed to near-zero variance.
\end{description}

\subsection{The Spectral Test for Kernel Schedule Analysis}
\label{sec:3.3.4}

The spectral test (TAOCP~\S3.3.4) is the most powerful theoretical test for detecting lattice structure in sequences generated by linear recurrences. In the kernel auto-tuning context, we apply it as follows.

Consider a kernel auto-tuner that generates a sequence of tiling configurations $\mathbf{x}_1, \mathbf{x}_2, \ldots \in \mathbb{Z}^d$ where $d$ is the number of tunable parameters (tile dimensions, VLIW slot assignments, prefetch depths). If the auto-tuner uses a deterministic update rule $\mathbf{x}_{n+1} = f(\mathbf{x}_n)$, and if $f$ has linear structure (e.g., it is a multi-dimensional linear congruential generator), then the configurations will lie on a lattice in $\mathbb{Z}^d$.

The spectral test computes the minimum distance $\nu_d$ between adjacent hyperplanes of this lattice. A small $\nu_d$ means the configurations are clustered on widely-spaced parallel hyperplanes---they fail to explore the configuration space uniformly. For the Trainium2 GEMM auto-tuner with $d=4$ parameters (M-tile, N-tile, K-tile, VLIW-pack), the spectral test reveals whether the tuner is visiting all viable tiling regions or is trapped in a low-dimensional subspace.

More importantly for agentic RL: the spectral test can be applied to the \emph{policy parameter trajectory} during training. If the gradient updates cause the policy parameters to trace a path that lies on a low-dimensional lattice in parameter space, the training has degenerated into a periodic orbit---the policy is oscillating between a small number of behaviors rather than converging. This is the high-dimensional analog of the cycle degeneration detected by Floyd's algorithm in Experiment~3.1A.

\subsection{Exploration Collapse Detection}
\label{sec:3.3.5}

Issue~\#490 identifies a structural failure mode that the validation tests above are designed to detect but that deserves explicit treatment: \emph{exploration collapse}, where an agent converges to a fixed policy and loses the ability to discover novel behaviors not represented in the reward signal. The issue documents five concrete engineering gaps in current agentic RL systems, each mappable to a validation test from this chapter.

\medskip
\noindent\textbf{(i) Trajectory Entropy Monitoring.} Define the \emph{trajectory entropy} over a sliding window of $W$ episodes as
\[
H_W = -\sum_{i=1}^{K} p_i \log p_i
\]
where $p_i$ is the empirical frequency of the $i$-th distinct tool-call sequence pattern in the window. If $H_W$ drops below a threshold $H_{\min}$ while the mean reward plateaus (i.e., $|\bar{R}_W - \bar{R}_{W-1}| < \epsilon$), the agent has collapsed to a fixed policy rather than mastering the task space. In Agent Lightning's architecture, the \texttt{tracer} records the action sequences needed to compute $H_W$; the \texttt{store} backends persist them across training steps. This metric is the agentic analog of the ``run test'' from \S\ref{sec:3.3.2}: a monotonically decreasing $H_W$ produces an excessively long ``run down'' in the entropy sequence.

\medskip
\noindent\textbf{(ii) Reward Staleness Detection.} The reward function is assumed correct for the duration of training, but in real deployments, environment dynamics shift---APIs evolve, data distributions drift, new capabilities appear. Issue~\#490 proposes tracking the correlation between the reward signal and an independently measured task success metric. When this correlation degrades below a threshold, the reward function should be flagged for review. In the integration bridge (\S\ref{sec:3.2.agenticrl}), the \texttt{PolicyReward} class from \texttt{operatorRL/src/agent\_os/integrations/agent\_lightning/reward.py} already maintains per-epoch statistics (\texttt{get\_stats()}); extending it to track reward--success correlation requires only a running Pearson coefficient over the \texttt{CompositeReward} components.

\medskip
\noindent\textbf{(iii) Novelty--Routine Classification.} Agent Lightning's \texttt{emit\_*()} functions capture what the agent did and what reward it received, but not whether the behavior was \emph{novel} (first occurrence) or \emph{routine} (repetition of a known-good policy). The four quadrants of the novelty--reward matrix carry different training implications:
\begin{center}
\begin{tabular}{@{}lll@{}}
\toprule
& High Reward & Low Reward \\
\midrule
Novel behavior & Discovery (study, don't just reinforce) & Exploration (valuable signal) \\
Routine behavior & Exploitation (fine, but not learning) & Degradation (flag immediately) \\
\bottomrule
\end{tabular}
\end{center}
Implementing novelty detection requires comparing each trajectory against the historical trace distribution---a computation that the \texttt{FlightRecorderEmitter}'s \texttt{get\_violation\_summary()} could be extended to support by hashing trajectory action sequences and maintaining a Bloom filter of seen patterns.

\subsection{Credit Assignment in Multi-Turn Agentic RL}
\label{sec:3.3.credit}

The validation tests above assume that each training signal (reward, gradient, utilization metric) can be attributed to a specific cause. In single-step RL, this attribution is straightforward: the reward for action $a_t$ in state $s_t$ informs the policy gradient for that state-action pair. In multi-turn agentic RL---the setting of Agent Lightning's \texttt{runner} and \texttt{tracer} components---the agent executes a sequence of tool-use actions $(a_1, a_2, \ldots, a_T)$ over an episode of length $T$, and a single scalar reward $R$ is observed only at the end. The credit assignment problem is: \emph{which actions contributed positively to $R$, and which were wasted or harmful?}

This problem is central to agentic RL and is discussed in Agent Lightning Issues~\#31, \#282, \#390, and \#392. Three credit assignment strategies are available, each with distinct hardware implications:

\begin{description}
\item[Trajectory-Level Credit.] The entire reward $R$ is attributed uniformly to all actions in the episode. The policy gradient for each action is $\nabla_\theta \log \pi_\theta(a_t | s_t) \cdot (R - b)$ where $b$ is a baseline. This is the simplest strategy and the one currently implemented in Agent Lightning's GRPO backend (via VERL). Its advantage is minimal memory overhead: only the scalar reward $R$ and the trajectory log-probabilities need to be stored. On Trainium2, the \texttt{trace\_aggregator} at trajectory level (\S\ref{sec:3.2.daemon}) suffices. Its disadvantage: high gradient variance, scaling as $O(T)$ with episode length, which slows convergence.

\item[Transition-Level Credit.] A reward model assigns intermediate rewards $r_t$ at each step. The policy gradient uses per-step advantages $\hat{A}_t = r_t + \gamma V(s_{t+1}) - V(s_t)$ (GAE-style). This reduces variance but requires: (a)~a reward model that can evaluate partial trajectories, and (b)~storing $T$ intermediate reward values plus the value function estimates---increasing the memory footprint by a factor of $\sim 3T$ relative to trajectory-level. On Trainium2, this means the XLA runtime's memory planner must accommodate the additional tensors, and for long episodes ($T > 100$), OOM becomes a real risk (Issue~\#107, \#438).

\item[Action-Level Credit.] Fine-grained credit assignment via token-level rewards (e.g., process reward models). Each \emph{token} in the agent's output receives a reward signal. This is the most informative but most expensive strategy: the memory footprint scales as $O(T \cdot L)$ where $L$ is the average response length in tokens. For the Agent Lightning + Operator RL system, action-level credit requires the \texttt{trace\_aggregator} to operate at transition level with per-token annotation---a mode not yet fully supported (see Issue~\#392 for the proposed taxonomy).
\end{description}

The choice of credit assignment strategy directly affects which validation tests from \S\ref{sec:3.3.1}--\S\ref{sec:3.3.4} are applicable. The serial test (\S\ref{sec:3.3.2}) requires transition-level data to compute consecutive action correlations. The poker test requires at minimum trajectory-level action sequences. The $\chi^2$ reward distribution test (\S\ref{sec:3.3.1}) is meaningful only with trajectory-level rewards unless intermediate rewards are available. Designing a validation suite thus requires committing to a credit assignment granularity first---a constraint that propagates all the way down to the hardware memory allocation decisions of \S\ref{sec:3.2.migration.memory}.

% ============================================================
\section{Exercises}
\label{sec:3.2.exercises}
% ============================================================

\begin{enumerate}

\item[\textbf{1.}] [20] Compare the memory hierarchy of an NVIDIA H100 GPU (HBM3 $\to$ L2 $\to$ shared memory $\to$ registers) with that of a Trainium2 chip (HBM $\to$ SBUF/PSUM scratchpad $\to$ engine-local registers). For a GEMM kernel with tile size $128 \times 128$ in BF16, compute the minimum data movement in bytes for each platform, assuming optimal tiling.

\item[\textbf{2.}] [$M20$] Megatron's tensor parallelism performs $2$ all-reduce operations per transformer layer in the forward pass and $2$ in the backward pass. If the model has $L$ layers, the hidden dimension is $H$, and the sequence length is $S$, express the total communication volume (in bytes, assuming BF16) as a function of $L$, $H$, $S$, and the tensor parallelism degree $T$. Under what conditions does the computation-to-communication ratio favor Trainium2's NeuronLink (800 GB/s bidirectional) over NVIDIA's NVLink (900 GB/s on H100)?

\item[\textbf{3.}] [$HM25$] The NKI Library provides optimized attention kernels in two variants: CTE (Compile-Time Execution) and TKG. Describe the algorithmic difference between these variants and explain why a single universal attention kernel cannot achieve peak utilization on Trainium2's VLIW architecture for all sequence lengths. (Hint: consider the relationship between the VLIW instruction bundle width and the number of independent operations available at different phases of the attention computation.)

\item[\textbf{4.}] [$HM35$] Design a hardware abstraction layer for the Operator RL system (\texttt{operatorRL/src/agent\_os}) that dispatches collective communication operations to NCCL (GPU), XLA collectives (TPU), or NxD (Trainium2) based on a runtime hardware detection mechanism. Your design should support the ``hot migration'' scenario where an agentic rollout worker is dynamically reassigned from one accelerator type to another mid-episode. Analyze the state synchronization overhead in terms of the number of bytes that must be transferred.

\item[\textbf{5.}] [$M22$] In the Agent Lightning training loop (\texttt{agentlightning/trainer}), the gradient synchronization step uses all-reduce across data-parallel workers. If the Trainium2 cluster has a 2-level topology (NeuronLink within a node, EFA across nodes), design a hierarchical all-reduce that minimizes the total communication time. Express the time in terms of the message size $M$, the number of NeuronCores per node $N$, the number of nodes $P$, the NeuronLink bandwidth $\beta_1$, and the EFA bandwidth $\beta_2$. Compare with the ring all-reduce algorithm used by NCCL on a flat NVLink topology.

\item[\textbf{6.}] [$M22$] (Relates to Chapter~3.1, Exercise~6) Consider a kernel auto-tuner that searches for configurations on Trainium2 using the recurrence $\mathbf{X}_{n+1} = f(\mathbf{X}_n)$ where $\mathbf{X}$ is a vector of tiling parameters and $f$ is a function that evaluates the kernel's utilization and proposes a new configuration. Using the theory of ultimately periodic sequences from Chapter~3.1, prove that the auto-tuner must eventually cycle, and design an algorithm (analogous to Floyd's or Brent's cycle detection) that detects this degeneration using $O(1)$ additional memory.

\item[\textbf{7.}] [$HM45$] Implement a complete, working NKI kernel for fused RMSNorm with BF16-to-FP8 quantization on Trainium2, and compare its throughput against the NKI Library's \texttt{nkilib.rmsnorm\_quant} kernel for hidden dimensions $H \in \{4096, 8192, 16384\}$. Discuss any performance gaps in terms of instruction scheduling efficiency (VLIW slot utilization) and memory access patterns (SBUF bank conflicts, if applicable). \textit{(Note: this exercise requires Trainium2 hardware access and familiarity with NKI \citep{awsneuron2025nki}. Expect several weeks of engineering effort.)}

\item[\textbf{8.}] [$HM30$] The companion document \emph{Anthropic~WantToDoWhat} states that Anthropic is ``co-designing next-generation Trainium hardware'' with Annapurna Labs. Formalize this hardware-software co-design as an optimization problem: given a set of critical kernel shapes $\mathcal{K} = \{k_1, \ldots, k_n\}$ and hardware parameters $\mathcal{H} = \{h_1, \ldots, h_m\}$, define an objective function that measures aggregate throughput. \textit{Explain why the resulting optimization is non-convex} (discrete hardware parameters, non-monotone tile-size/pipeline interactions, integer constraints on engine counts). Propose a relaxation or heuristic approach.

\item[\textbf{9.}] [$HM30$] \textbf{(Credit Assignment)} Design a credit assignment algorithm for Agent Lightning that assigns rewards to individual tool-use steps in a multi-turn rollout of length $T$. Derive the variance of the advantage estimate $\hat{A}_t$ under: (a)~trajectory-level credit ($\hat{A}_t = R - b$ for all $t$), (b)~transition-level credit ($\hat{A}_t = r_t + \gamma V(s_{t+1}) - V(s_t)$), and (c)~action-level credit (per-token reward). Express each variance as a function of $T$ and the reward signal noise $\sigma^2_r$. Under what conditions does transition-level credit outweigh its additional memory cost on Trainium2? (See \S\ref{sec:3.3.credit}.)

\item[\textbf{10.}] [$HM50$] \textbf{(Open Research Problem.)} Issue~\#490 reports exploration collapse: agents converge to fixed policies. Formalize this as cycle degeneration in policy parameter space (cf.\ \S\ref{sec:3.1}). Propose a detection mechanism based on the spectral test (\S\ref{sec:3.3.4}) applied to the policy gradient trajectory, and design an intervention that forces the policy off its periodic orbit without destroying learned behavior. Analyze the hardware cost on Trainium2.

\item[\textbf{11.}] [$M20$] \textbf{(From Issue~\#489: GRPO Grouping.)} In multi-turn agent RL, GRPO normalizes advantages within a group of $G$ rollouts that share the same prompt. Issue~\#489 asks: is it valid to mix samples from \emph{different} prompts in the same GRPO group? Consider the following concrete example (from the \texttt{calc\_x} agent with \texttt{rollout\_n=4}):

\begin{center}
\begin{tabular}{@{}llllr@{}}
\toprule
Sample & Rollout & Turn & Prompt Context & Reward \\
\midrule
1 & Rollout~1, Turn~0 & \texttt{[system, user]} & Initial query & 1.0 \\
2 & Rollout~1, Turn~1 & \texttt{[..., assistant, tool\_response]} & After tool call & 1.0 \\
3 & Rollout~2, Turn~0 & \texttt{[system, user]} & Initial query (no tool) & 0.0 \\
4 & Rollout~3, Turn~0 & \texttt{[system, user]} & Initial query & 1.0 \\
5 & Rollout~3, Turn~1 & \texttt{[..., assistant, tool\_response]} & After tool call & 1.0 \\
6 & Rollout~4, Turn~0 & \texttt{[system, user]} & Initial query (no tool) & 0.0 \\
\bottomrule
\end{tabular}
\end{center}

All 6 samples share the same \texttt{uid} and are grouped together for GRPO advantage computation. But Turn~0 samples have prompt $p_0 = [\text{system}, \text{user}]$, while Turn~1 samples have prompt $p_1 = [\text{system}, \text{user}, \text{assistant}, \text{tool\_response}]$---fundamentally different inputs. (a)~Prove that if prompts $p_0, p_1$ have reward distributions $R_0 \sim \mathcal{N}(\mu_0, \sigma_0^2)$ and $R_1 \sim \mathcal{N}(\mu_1, \sigma_1^2)$ with $\mu_0 \neq \mu_1$, then the group-normalized advantage $\hat{A}_i = (R_i - \bar{R}) / \sigma_R$ is \emph{biased}: actions under prompt $p_0$ receive systematically higher (or lower) advantages regardless of their quality. (b)~Derive the bias magnitude as a function of $|\mu_0 - \mu_1|$, $G$, and the mixing ratio $n_0/n_1$. (c)~In the example above, Turn~0 samples with $R=0$ (no tool use) receive negative advantages, while Turn~1 samples (post-tool-call) receive positive advantages. Explain why this creates a gradient signal that \emph{discourages tool use at Turn~0}---the opposite of the intended behavior---and propose a correction. (Hint: consider computing advantages separately within each prompt group.)

\item[\textbf{12.}] [$HM25$] \textbf{(From Issue~\#470: Local Models.)} Agent Lightning's daemon (\texttt{agentlightning/verl/daemon.py}) uses the \texttt{\_update\_proxy\_server\_v1()} function to route LLM inference requests. Modify this function to support a local vLLM \citep{kwon2023vllm} server running on Trainium2 via NxD Inference. Specifically: (a)~identify which HTTP endpoints the proxy must implement, (b)~explain why the \texttt{/v1/completions} endpoint's token streaming behavior differs between CUDA-backed vLLM and Neuron-backed vLLM (hint: Neuron's pipeline execution model produces tokens in batches rather than one-by-one), and (c)~propose a buffering strategy that masks this difference from the agent.

\item[\textbf{13.}] [$M25$] \textbf{(From Issue~\#448: Failed Rollouts.)} When an agent rollout fails mid-episode (e.g., tool-use timeout, environment crash), the \texttt{trace\_aggregator} receives an incomplete trajectory of length $T' < T$. The commit \texttt{19c72db6} handles empty tensors to prevent crashes, but the RL algorithm must still decide how to use partial data. Define a formal framework: let $\tau_\text{partial} = (s_1, a_1, r_1, \ldots, s_{T'}, a_{T'})$ be an incomplete trajectory truncated by failure. (a)~Under trajectory-level credit, show that na\"ively assigning $R = 0$ introduces a systematic negative bias of magnitude $\mathbb{E}[R | \text{complete}] \cdot P(\text{failure})$. (b)~Propose a correction using importance weighting. (c)~Analyze the memory overhead of storing partial trajectories on Trainium2, given that the XLA memory planner pre-allocates for full episode length $T$.

\end{enumerate}

% ============================================================

% ============================================================
\appendix
\section{Operator RL Module Reference}
\label{app:operatorrl-modules}
% ============================================================

The Operator RL repository (\texttt{operatorRL}) organizes its kernel architecture into the following module hierarchy. Each module corresponds to one or more layers of the 4-layer agent governance stack described in \S\ref{sec:3.2.agenticrl}.

\begin{description}
\item[\texttt{modules/primitives}] \textbf{(Layer~1: Core Primitives.)} Fundamental operations for agent execution: action dispatch, observation parsing, reward signal routing. These are the lowest-level building blocks, analogous to TAOCP's MIX instruction set --- the atomic operations from which all higher-level agent behavior is composed.

\item[\texttt{modules/iatp}] \textbf{(Layer~2: Inter-Agent Trust Protocol.)} Manages trust boundaries between cooperating agents. When agents share hardware resources (e.g., multiple agents training on the same Trainium2 node), IATP enforces isolation guarantees: one agent's policy update cannot corrupt another's gradient buffer. The \texttt{SIGTRUST} signal (\S\ref{sec:3.2.agenticrl}, Table~\ref{tab:signals}) is IATP's enforcement mechanism.

\item[\texttt{modules/emk}] \textbf{(Episodic Memory Kernel.)} Manages episodic memory across multi-turn rollouts. In agentic RL, the agent must remember past actions within an episode (working memory) and across episodes (episodic memory). EMK provides a kernel-level API for storing and retrieving episodic context, backed by the VFS mount at \texttt{/agent/\{id\}/mem/episodic}.

\item[\texttt{modules/scak}] \textbf{(Layer~4: Self-Correcting Agent Kernel.)} The highest layer: detects policy degradation, triggers self-correction, and manages the feedback loop between the agent's runtime behavior and its training updates. SCAK consumes the validation test results from \S\ref{sec:3.3} --- if the $\chi^2$ utilization test or the spectral test detects anomalies, SCAK can issue a retraining signal.

\item[\texttt{modules/cmvk}] \textbf{(Cross-Model Verification Kernel.)} Ensures numerical consistency when the same model runs on different hardware backends. After migrating a kernel from CUDA to NKI (\S\ref{sec:3.2.migration.kernels}), CMVK runs a verification suite comparing output tensors within a specified tolerance ($\epsilon \leq 10^{-3}$ for BF16 operations).

\item[\texttt{modules/nexus}] \textbf{(Multi-Agent Coordination.)} Orchestrates multiple agents operating concurrently. Nexus manages agent lifecycle (spawn, pause, resume, terminate), resource allocation across NeuronCores, and inter-agent message passing. Issue~\#92 in Agent Lightning asks about multi-agent support; Nexus is Operator RL's answer.

\item[\texttt{modules/amb}] \textbf{(Agent Memory Buffer.)} A ring buffer for short-term agent state, used during rollout execution. AMB is the runtime complement to EMK's persistent episodic storage.

\item[\texttt{modules/caas}] \textbf{(Compliance-as-a-Service.)} Governance module for regulatory compliance in self-deploying agents. CaaS logs all agent actions to an immutable audit trail and enforces domain-specific constraints (e.g., financial transaction limits, PII handling policies).

\item[\texttt{modules/observability}] \textbf{(Observability.)} Metrics, traces, and logs for the kernel itself. Integrates with Agent Lightning's dashboard (\texttt{agentlightning/dashboard}) and the Flight Recorder.

\item[\texttt{benchmarks/bench\_kernel.py}] Benchmarks for kernel-level operations. See Exercise~4 (\S\ref{sec:3.2.exercises}).

\item[\texttt{benchmarks/bench\_policy.py}] Benchmarks for the policy engine's evaluation latency --- critical for ensuring governance overhead remains negligible during training.

\item[\texttt{examples/self-evaluating}] A complete example of a self-evaluating agent that runs the validation tests from \S\ref{sec:3.3} on its own training metrics and triggers SCAK-mediated self-correction when anomalies are detected.
\end{description}

\section{Agent Lightning Component Reference}
\label{app:agentlightning-components}

\begin{description}
\item[\texttt{agentlightning/store/}] Seven backend implementations (memory, SQLite, MongoDB, client-server, collection-based, threaded, base) for the LightningStore coordination layer. The multiple backends mirror the modulus choice discussion in \S\ref{sec:3.2.1.1}: just as different moduli $m$ suit different hardware word sizes, different store backends suit different deployment scales.

\item[\texttt{agentlightning/tracer/}] Captures the agent's interaction trace during rollout. The tracer records state-action-reward-next\_state tuples at the transition level and assembles them into trajectories. This data feeds both the training loop (via \texttt{trace\_aggregator}) and the validation tests of \S\ref{sec:3.3}.

\item[\texttt{agentlightning/adapter/}] Framework adapters (LangChain, CrewAI, etc.) that convert diverse agent framework outputs into the standardized trace format. The \texttt{TracerTraceToTriplet} adapter family handles the conversion from raw traces to (state, action, reward) triplets suitable for RL training.

\item[\texttt{agentlightning/algorithm/}] RL algorithm implementations. As of early 2026, supports GRPO and PPO via the VERL backend. Each algorithm defines its own advantage estimation and gradient computation, which determines the hardware memory requirements during the policy update step (\S\ref{sec:3.2.migration.memory}).
\end{description}

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}